---
title: "COVID-19 County-Level Mortality Analysis"
subtitle: "Excess Death Rates, Demographic Drivers, Seasonal Variance and Outlier Detection"
author: "Brian E. Ketchens"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 12
    fig_height: 7
---

```{css, echo=FALSE}
/* Fix TOC wrapped line alignment */
.tocify-item {
  white-space: normal !important;
}

.tocify ul li a {
  display: block;
  padding-left: 10px;
  text-indent: -10px;
  padding-right: 8px;
}

.tocify-subheader .tocify-item a {
  padding-left: 25px;
  text-indent: -10px;
}

.tocify-subheader .tocify-subheader .tocify-item a {
  padding-left: 40px;
  text-indent: -10px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  cache     = FALSE
)
```



# 1. Environment & Data Connection

```{r libraries}
# ---- Core ----
library(tidyverse)
library(lubridate)
library(readr)

# ---- BigQuery connection (retained for Option 3 development) ----
library(bigrquery)
library(DBI)

# ---- Statistical modeling ----
library(MASS)
library(lme4)
library(broom)
library(broom.mixed)

# ---- Time series / seasonal decomposition ----
library(forecast)
library(tsibble)
library(feasts)

# ---- Visualization ----
library(scales)
library(ggrepel)
library(patchwork)
library(viridis)
library(RColorBrewer)
library(ggridges)

# ---- Spatial ----
library(tigris)
library(sf)

# ---- Tables ----
library(gt)
library(kableExtra)

# ---- Hexbin ---
install.packages("hexbin")

# ---- Plotly --- 
install.packages("plotly")

# ---- Leaflet ---
install.packages("leaflet")
```

```{r tigris_setup, include=FALSE, message=FALSE, warning=FALSE}
# ============================================================
# Pre-download and cache all Census shapefiles silently
# Prevents tigris progress bars from appearing in knit output
# After first run shapefiles are cached and this is instant
# ============================================================
options(tigris_use_cache = TRUE)

suppressMessages(
  suppressWarnings({
    counties_sf <- counties(cb = TRUE, resolution = "20m", 
                            year = 2020) %>%
      mutate(fips = paste0(STATEFP, COUNTYFP)) %>%
      filter(!STATEFP %in% c("02", "15", "60", 
                              "66", "69", "72", "78"))
    
    all_us_counties <- counties_sf %>%
      as.data.frame() %>%
      dplyr::select(fips, STATEFP, NAME, STATE_NAME)
  })
)
```

```{r bq_connection}
# BigQuery live connection not needed for CSV workflow.
# Retained for reference when transitioning to state-level
# development queries in Option 3.

# PROJECT_ID <- "covid-explorer"
# bq_auth(path = NULL)
# con <- dbConnect(
#   bigrquery::bigquery(),
#   project = PROJECT_ID,
#   dataset = "covid_analytics",
#   billing = PROJECT_ID
# )
```

```{r load_data, results ='hide', message=FALSE}
# ============================================================
# Load county_weekly from CSV
# Negative value clamping and weekly aggregation were performed
# in BigQuery before export -- no raw daily data needed here.
# ============================================================

county_weekly <- read_csv("county_weekly.csv") %>%
  mutate(
    week_start    = as.Date(week_start),
    fips          = as.character(fips),
    pandemic_wave = factor(pandemic_wave, levels = c(
      "pre_pandemic", "1_initial", "2_summer2020", "3_winter2021",
      "4_alpha", "5_delta", "6_omicron", "7_endemic"
    )),
    vaccine_era = factor(vaccine_era, levels = c(
      "pre_vaccine", "early_rollout", "broad_availability"
    )),
    year     = as.integer(year),
    month    = as.integer(month),
    iso_week = as.integer(iso_week)
  )

cat("Rows loaded:  ", nrow(county_weekly), "\n")
cat("Counties:     ", n_distinct(county_weekly$fips), "\n")
cat("Date range:   ", as.character(min(county_weekly$week_start)),
    "to", as.character(max(county_weekly$week_start)), "\n")
```



# 2. Data Quality & Exploratory Overview

```{r data_quality}
missing_summary <- county_weekly %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") %>%
  filter(pct_missing > 0) %>%
  arrange(desc(pct_missing))

missing_summary %>%
  mutate(pct_missing = scales::percent(pct_missing, accuracy = 0.1)) %>%
  kable(caption = "Columns with Missing Values") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
```{r data_cleanup}
# Drop columns with 100% missingness -- no analytical value
county_weekly <- county_weekly %>%
  dplyr::select(-new_tested_wk, -test_positivity_wk)

# Mobility coverage note
mobility_coverage <- county_weekly %>%
  summarise(
    counties_with_mobility = n_distinct(fips[!is.na(mobility_retail)]),
    total_counties         = n_distinct(fips),
    pct_coverage           = scales::percent(
      n_distinct(fips[!is.na(mobility_retail)]) / n_distinct(fips),
      accuracy = 0.1
    )
  )

cat("Counties with mobility data:", mobility_coverage$counties_with_mobility, 
    "of", mobility_coverage$total_counties,
    "(", mobility_coverage$pct_coverage, ")\n")

# NOTE: Missing mobility data skews toward rural/small counties.
# Mobility variables will be used descriptively, not as model covariates.
# Treat NULL case and death counts as zero
# These represent weeks with no reported activity,
# not truly missing data -- the pipeline used NULL instead of 0
county_weekly <- county_weekly %>%
  mutate(
    new_confirmed_wk = replace_na(new_confirmed_wk, 0),
    new_deceased_wk  = replace_na(new_deceased_wk,  0),
    case_rate_wk     = replace_na(case_rate_wk,     0),
    death_rate_wk    = replace_na(death_rate_wk,    0),
    # Cumulative columns: carry forward last known value within each county
    # For now replace with 0 -- use max() in county_totals
    # to handle this correctly
    cumulative_confirmed = replace_na(cumulative_confirmed, 0),
    cumulative_deceased  = replace_na(cumulative_deceased,  0),
    cum_case_rate_100k   = replace_na(cum_case_rate_100k,   0),
    cum_death_rate_100k  = replace_na(cum_death_rate_100k,  0)
  )
```



# 3. National Overview


```{r national_trend, fig.cap="Weekly national COVID-19 cases and deaths per 100k population"}
national_weekly <- county_weekly %>%
  group_by(week_start, pandemic_wave) %>%
  summarise(
    case_rate_national  = weighted.mean(case_rate_wk,  total_pop, na.rm = TRUE),
    death_rate_national = weighted.mean(death_rate_wk, total_pop, na.rm = TRUE),
    .groups = "drop"
  )

wave_colors <- c(
  "1_initial"    = "#E63946",
  "2_summer2020" = "#F4A261",
  "3_winter2021" = "#457B9D",
  "4_alpha"      = "#2A9D8F",
  "5_delta"      = "#E76F51",
  "6_omicron"    = "#6A0572",
  "7_endemic"    = "#A8DADC",
  "pre-pandemic" = "#CCCCCC"
)

p_cases <- ggplot(national_weekly,
                  aes(x = week_start, y = case_rate_national, fill = pandemic_wave)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = wave_colors, name = "Wave") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = comma) +
  labs(title = "Weekly Case Rate per 100k (Population-Weighted National Average)",
       x = NULL, y = "Cases per 100k") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

p_deaths <- ggplot(national_weekly,
                   aes(x = week_start, y = death_rate_national, fill = pandemic_wave)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = wave_colors, name = "Wave") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = comma) +
  labs(title = "Weekly Death Rate per 100k (Population-Weighted National Average)",
       x = NULL, y = "Deaths per 100k") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

p_cases / p_deaths
```


# 3b. Hypothesis Testing

The national trend analysis in Section 3 established the broad 
arc of the pandemic. Three specific hypotheses emerge from that 
picture that the data can formally test. Each hypothesis is 
stated, tested with an appropriate statistical method, and 
interpreted in the context of the structural inequity findings 
that follow in Sections 4 through 7.

```{r hypothesis_setup}
# ============================================================
# HYPOTHESIS TESTING SETUP
# Derive wave-specific county summaries for all three tests
# ============================================================

wave_county_rates <- county_weekly %>%
  filter(pandemic_wave != "pre_pandemic") %>%
  group_by(fips, county_name, state_name, pandemic_wave, total_pop) %>%
  summarise(
    wave_death_rate = (sum(new_deceased_wk,  na.rm = TRUE) / 
                       first(total_pop)) * 100000,
    wave_case_rate  = (sum(new_confirmed_wk, na.rm = TRUE) / 
                       first(total_pop)) * 100000,
    wave_cfr        = sum(new_deceased_wk,  na.rm = TRUE) /
                      pmax(sum(new_confirmed_wk, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  filter(wave_death_rate >= 0, wave_case_rate >= 0)

```

### Hypothesis 1: Did Prior Exposure Protect Counties During Delta?

**Stated Hypothesis:** Counties with higher vaccination rates had 
statistically lower death rates during the Delta wave after 
controlling for age structure and socioeconomic status.

**Methodological Note:** County-level vaccination data is not yet 
integrated into this dataset and will be added in Phase 2. In its 
absence this hypothesis is tested through a structural proxy — 
comparing each county's Wave 1 death rate against its Delta wave 
death rate. Counties that bore catastrophic Wave 1 burden had 
large portions of their populations exposed to the virus before 
vaccines were available. If prior infection conferred meaningful 
population-level protection, those same counties should show 
relative improvement during Delta regardless of vaccination 
coverage. This is a testable prediction the data can evaluate 
directly.

> **Why This Matters:** If prior infection immunity and vaccination 
> immunity produced similar protective effects at the population 
> level, the policy implications differ from a world where 
> vaccination alone drove Delta protection. The Phase 2 
> vaccination data integration will formally partition these 
> two mechanisms. For now the proxy approach tells us whether 
> the protection pattern is consistent with prior exposure 
> as a plausible explanation.

The paired t-test results and county-level scatter plot below 
examine whether Wave 1 burden predicted Delta improvement — 
and the answer is striking.

```{r hypothesis_1, fig.cap="H1: Did county vaccination rate predict lower Delta death rates?"}

delta_vs_wave1 <- wave_county_rates %>%
  filter(pandemic_wave %in% c("1_initial", "5_delta")) %>%
  dplyr::select(fips, county_name, state_name, pandemic_wave,
                wave_death_rate, total_pop) %>%
  pivot_wider(
    names_from  = pandemic_wave,
    values_from = wave_death_rate,
    names_prefix = "dr_"
  ) %>%
  left_join(
    county_weekly %>%
      dplyr::select(fips, pct_65_over, poverty_rate,
                    pct_black, pct_hispanic, median_age) %>%
      distinct(fips, .keep_all = TRUE),
    by = "fips"
  ) %>%
    
  drop_na()

# Paired t-test: did death rates differ between Wave 1 and Delta?
t_result <- t.test(delta_vs_wave1$dr_5_delta,
                   delta_vs_wave1$dr_1_initial,
                   paired = TRUE)

cat("Mean Wave 1 death rate:  ", round(mean(delta_vs_wave1$dr_1_initial, na.rm=TRUE), 2), "\n")
cat("Mean Delta death rate:   ", round(mean(delta_vs_wave1$dr_5_delta,   na.rm=TRUE), 2), "\n")
cat("Mean difference:         ", round(t_result$estimate, 2), "\n")
cat("95% CI:                  ", round(t_result$conf.int[1], 2),
    "to", round(t_result$conf.int[2], 2), "\n")
cat("p-value:                 ", scales::pvalue(t_result$p.value), "\n")

# Visual: scatter of Wave 1 vs Delta death rate by county
# Counties above the diagonal had WORSE Delta outcomes than Wave 1
delta_vs_wave1 %>%
  mutate(
    burden_shift = case_when(
      dr_5_delta > dr_1_initial * 1.5 ~ "Much worse in Delta",
      dr_5_delta > dr_1_initial       ~ "Worse in Delta",
      dr_5_delta < dr_1_initial * 0.5 ~ "Much better in Delta",
      TRUE                             ~ "Similar"
    ),
    burden_shift = factor(burden_shift, levels = c(
      "Much worse in Delta", "Worse in Delta",
      "Similar", "Much better in Delta"
    ))
  ) %>%
  ggplot(aes(x = dr_1_initial, y = dr_5_delta, color = burden_shift)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  geom_smooth(method = "lm", se = TRUE,
              color = "black", linewidth = 0.8) +
  scale_color_manual(
    values = c(
      "Much worse in Delta"  = "#E63946",
      "Worse in Delta"       = "#F4A261",
      "Similar"              = "grey60",
      "Much better in Delta" = "#457B9D"
    ),
    name = NULL
  ) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotate("text", x = Inf, y = -Inf,
           label = "Above line = worse in Delta than Wave 1",
           hjust = 1.1, vjust = -0.5, size = 3.5, color = "grey40") +
  labs(
    title    = "H1: County Death Rate - Wave 1 vs Delta",
    subtitle = "Each point = one county | Dashed line = equal burden",
    x        = "Wave 1 Death Rate per 100k",
    y        = "Delta Wave Death Rate per 100k",
    caption  = "NOTE: Vaccination rate covariate to be added when CDC data integrated"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Reading the H1 Results

The mean Delta wave death rate (78.6 per 100k) was dramatically 
higher than the mean Wave 1 death rate (17.1 per 100k) — a 
difference of 61.5 deaths per 100k that is statistically 
unambiguous (95% CI: 59.0 to 64.0, p<0.001). Delta was 
substantially more lethal at the county level than the initial 
outbreak by almost every measure.

However the scatter plot reveals something the aggregate numbers 
obscure. The 137 counties classified as Much Better in Delta — 
those that improved most dramatically from Wave 1 — cluster into 
three geographically and mechanistically distinct groups, each 
pointing toward a different protective pathway.

```{r h1_county_detail}
h1_detail <- delta_vs_wave1 %>%
  mutate(
    burden_shift = case_when(
      dr_5_delta > dr_1_initial * 1.5 ~ "Much worse in Delta",
      dr_5_delta > dr_1_initial       ~ "Worse in Delta",
      dr_5_delta < dr_1_initial * 0.5 ~ "Much better in Delta",
      TRUE                             ~ "Similar"
    ),
    burden_shift = factor(burden_shift, levels = c(
      "Much worse in Delta", "Worse in Delta",
      "Similar", "Much better in Delta"
    )),
    dr_1_initial = round(dr_1_initial, 1),
    dr_5_delta   = round(dr_5_delta,   1),
    delta_vs_w1  = round(dr_5_delta - dr_1_initial, 1)
  ) %>%
  dplyr::select(county_name, state_name, dr_1_initial,
                dr_5_delta, delta_vs_w1, burden_shift) %>%
  rename(
    County        = county_name,
    State         = state_name,
    Wave1_rate    = dr_1_initial,
    Delta_rate    = dr_5_delta,
    Difference    = delta_vs_w1,
    Burden_category = burden_shift
  ) %>%
  arrange(desc(Delta_rate))

# Summary count by category -- uses Burden_category
h1_detail %>%
  count(Burden_category) %>%
  mutate(pct = scales::percent(n / sum(n), accuracy = 0.1)) %>%
  print()

# Display table -- drop Burden_category for clean presentation
h1_detail %>%
  slice_head(n = 20) %>%
  dplyr::select(County, State, Wave1_rate,
                Delta_rate, Difference) %>%
  kable(caption = "Worst 20 Counties by Delta Wave Death Rate per 100k") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r h1_county_by_category}

cat("Counties classified as Much Better in Delta than Wave 1:\n")
h1_detail %>%
  filter(Burden_category == "Much better in Delta") %>%
  arrange(desc(Wave1_rate)) %>%
  slice_head(n = 30) %>% 
  dplyr::select(County, State, Wave1_rate, 
                Delta_rate, Difference) %>%   # Burden_category removed
  kable(caption = "Top 30 Counties That Improved: Highest Wave 1 Burden, Lower Delta Burden") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Protected Clusters

There were 137 counties classified as having substantially lower 
Delta wave death rates relative to their Wave 1 burden — 
counties where prior exposure appears to have conferred meaningful 
population-level protection against subsequent waves. The counties were classified into mechanistic subgroups for separate analytical treatment.

> **Analytical Caution:** The Deep South improvement magnitude 
> may be partially artifactual. Wave 1 death rates in these counties
> may be inflated by:
>
> 1. **Denominator problems** -- small county populations amplify rates.
> 2. **Attribution differences** -- rural hospitals may have coded COVID
>    deaths differently than urban centers.
> 3. **Reporting lag** -- smaller counties sometimes batch-reported deaths.
>
> These hypotheses cannot be resolved without death certificate
> microdata but should be noted as limitations in the findings.

```{r h1_protected_clusters}

protected_counties <- h1_detail %>%
  filter(Burden_category == "Much better in Delta") %>%
  mutate(
    protection_hypothesis = case_when(
      # Northeast urban corridor -- vaccination + prior immunity
      State %in% c("New York", "New Jersey", "Massachusetts",
                   "Connecticut", "Rhode Island") &
        Wave1_rate > 50                          ~ "Urban NE: Vaccination + Prior Immunity",

      # Deep South Black Belt -- prior infection, low vaccination
      State %in% c("Georgia", "Mississippi",
                   "Louisiana", "Alabama") &
        Wave1_rate > 100                         ~ "Deep South: Prior Infection",

      # Nebraska meatpacking cluster
      State == "Nebraska"                        ~ "NEB Meatpacking: Occupational Prior Immunity",

      # Mid-Atlantic urban
      State %in% c("Maryland", "Pennsylvania",
                   "Virginia", "District of Columbia") ~ "Mid-Atlantic Urban",

      # Everything else
      TRUE                                       ~ "Other"
    )
  )

# Summary by hypothesis group
protected_counties %>%
  group_by(protection_hypothesis) %>%
  summarise(
    n_counties       = n(),
    mean_wave1_rate  = round(mean(Wave1_rate, na.rm = TRUE), 1),
    mean_delta_rate  = round(mean(Delta_rate, na.rm = TRUE), 1),
    mean_improvement = round(mean(abs(Difference), na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_improvement)) %>%
  kable(caption = "Protected County Clusters: Three Distinct Protective Mechanisms") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Three Distinct Protective Mechanisms

The 137 counties that resisted Delta's surge fall into three 
analytically distinct clusters that tell different causal stories:

**Urban Northeast — Vaccination and Prior Immunity (48 counties)**
New York, New Jersey, Massachusetts, and neighboring states 
dominated this cluster with a mean improvement of 115.6 deaths 
per 100k. These densely populated counties were devastated in 
Wave 1 and then dramatically protected during Delta — almost 
certainly reflecting the combined effect of high vaccination 
rates and widespread prior infection creating a high threshold 
of population immunity before Delta arrived.

**Deep South — Prior Infection Without Vaccination (14 counties)**
This is the most analytically striking cluster. Fourteen small 
rural majority Black counties in Georgia, Mississippi, and 
Louisiana achieved the greatest absolute improvement of any 
group — a mean of 136.1 deaths per 100k — despite almost 
certainly having lower vaccination rates than the Urban Northeast 
cluster. Their catastrophic Wave 1 burden, likely substantially 
undercounted due to limited rural testing infrastructure, appears 
to have created population-level immunity that persisted into the 
Delta wave. The data cannot confirm this mechanism without 
vaccination coverage data but it is the most parsimonious 
explanation consistent with the pattern observed.

**Nebraska Meatpacking Counties — Occupational Prior Immunity 
(11 counties)**
Dakota, Hall, Hamilton, Colfax, and neighboring Nebraska counties 
with large meatpacking plant workforces experienced explosive 
outbreaks in Wave 1 when plants became superspreader environments. 
By Delta their mean death rate fell to just 4.5 per 100k — a 92% 
reduction from Wave 1. This is a textbook occupational prior 
immunity cluster and one of the most analytically clean natural 
experiments in the dataset.

> **Analytical Caution:** The ordering of improvement magnitude — 
> Deep South counties improving more than Urban Northeast counties 
> despite likely lower vaccination coverage — is consistent with 
> prior infection immunity being at least as protective as 
> vaccination at the population level during the Delta wave. 
> This finding should be interpreted carefully and will be 
> formally tested when vaccination data is integrated in Phase 2. 
> It does not argue against vaccination — it argues that the 
> mechanisms of population protection are multiple and that 
> prior infection contributed meaningfully to the geographic 
> reorganization of burden documented in Hypothesis 3.

### Hypothesis 2: Did COVID Cases Become Less Lethal After Omicron?

**Stated Hypothesis:** The structural relationship between case 
rate and death rate changed after Omicron, reflecting some 
combination of variant attenuation, accumulated population 
immunity, and changes in death coding practices.

To test this, separate linear regressions were fitted relating 
weekly case rates to weekly death rates for the pre-Omicron 
period and the Omicron and post-Omicron period. If cases became 
less lethal the slope of the post-Omicron regression line should 
be meaningfully flatter than the pre-Omicron slope — meaning 
each additional case produced fewer deaths.

```{r hypothesis_2, fig.cap="H2: Did the case-to-death relationship change post-Omicron?"}
# ============================================================
# Method: fit separate OLS regressions of death_rate ~ case_rate
# for pre-Omicron vs post-Omicron periods and compare slopes.
# ============================================================

cfr_structure <- county_weekly %>%
  filter(new_confirmed_wk > 0, new_deceased_wk >= 0) %>%
  mutate(
    era = case_when(
      pandemic_wave %in% c("1_initial", "2_summer2020",
                           "3_winter2021", "4_alpha",
                           "5_delta")           ~ "Pre-Omicron",
      pandemic_wave %in% c("6_omicron",
                           "7_endemic")         ~ "Omicron & Post",
      TRUE                                       ~ NA_character_
    )
  ) %>%
  filter(!is.na(era))

# Fit separate models per era
models_by_era <- cfr_structure %>%
  group_by(era) %>%
  summarise(
    n           = n(),
    slope       = coef(lm(death_rate_wk ~ case_rate_wk))[2],
    intercept   = coef(lm(death_rate_wk ~ case_rate_wk))[1],
    r_squared   = summary(lm(death_rate_wk ~ case_rate_wk))$r.squared,
    .groups = "drop"
  )

models_by_era %>%
  mutate(across(c(slope, intercept, r_squared), ~round(., 4))) %>%
  print()

# Two clean regression lines on one plot -- no points
# The slope difference IS the finding

ggplot(cfr_structure %>%
         filter(case_rate_wk < 300, death_rate_wk < 15),
       aes(x = case_rate_wk, y = death_rate_wk, color = era)) +
  geom_smooth(
    method    = "lm",
    se        = TRUE,
    linewidth = 2.0,
    alpha     = 0.2      # confidence band transparency
  ) +
  scale_color_manual(
    values = c("Pre-Omicron"    = "#E63946",
               "Omicron & Post" = "#1D6FA4"),
    name   = NULL
  ) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotate("text",
           x = 200, y = 11,
           label = "Pre-Omicron: 0.0089 deaths per case",
           color = "#E63946", size = 4.5, fontface = "bold", hjust = 0) +
  annotate("text",
           x = 200, y = 9.5,
           label = "Omicron & Post: 0.0013 deaths per case",
           color = "#1D6FA4", size = 4.5, fontface = "bold", hjust = 0) +
  annotate("text",
           x = 200, y = 8.0,
           label = "85% reduction in case lethality",
           color = "grey30", size = 4.0, fontface = "italic", hjust = 0) +
  labs(
    title    = "H2: Did COVID Cases Become Less Lethal Post-Omicron?",
    subtitle = "Regression lines fitted to 300k county-weeks | Bands = 95% confidence interval",
    x        = "Weekly Case Rate per 100k",
    y        = "Weekly Death Rate per 100k",
    caption  = "Pre-Omicron slope 6.8x steeper than Omicron & Post -- consistent with attenuation and/or immunity"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Reading the H2 Results

The evidence is unambiguous. The slope of the case rate to death 
rate relationship dropped by approximately 85% between the 
pre-Omicron period (slope = 0.0089) and the Omicron and 
post-Omicron period (slope = 0.0013). A case of COVID in the 
endemic era was approximately seven times less likely to result 
in death than a case during the Delta wave or earlier.

The chart makes this visible without any statistical training 
required — the red pre-Omicron line rises steeply as case rates 
increase while the blue post-Omicron line is nearly flat across 
the same range.

Three mechanisms likely contributed to this structural change 
and the data cannot cleanly separate them:

**Variant attenuation** — Omicron and subsequent variants were 
intrinsically less virulent at the individual level, a finding 
supported by clinical literature independently of this analysis.

**Population immunity** — by late 2021 a substantial fraction 
of the US population carried immunity from vaccination, prior 
infection, or both, reducing the probability of severe disease 
per exposure even when cases were high.

**Surveillance and coding changes** — the elevated intercept 
in the post-Omicron model (3.31 vs 1.76 deaths per 100k at 
zero cases) suggests deaths were occurring in the endemic phase 
that were not being linked to active case counts. This is 
consistent with documented changes in how COVID deaths were 
attributed on death certificates beginning in 2022.

> **What the R-squared Values Tell Us:** Both models show low 
> R-squared values — 0.062 pre-Omicron and 0.009 post-Omicron. 
> This means case rate alone explains very little of the 
> variation in death rate at the county level in either period. 
> County-level structural factors — the demographic and 
> socioeconomic characteristics examined throughout this 
> analysis — explain far more of the variation than raw case 
> counts. This validates the county-level analytical approach 
> of this entire document.


### Hypothesis 3: Did the Same Counties Bear the Burden 
Throughout the Pandemic?

**Stated Hypothesis:** Counties with the highest Wave 1 death 
rates were not necessarily the same counties that drove 
subsequent waves. Pandemic burden shifted geographically as 
variants changed and population immunity accumulated.

Spearman rank correlation measures whether the ordering of 
counties by death rate stayed consistent across waves. A 
correlation near 1.0 means the same counties were consistently 
hardest hit — the burden was stable and predictable. A 
correlation near zero means knowing a county's rank in one 
wave tells you nothing about its rank in another — the burden 
reorganized completely. A negative correlation would mean 
counties that fared worst in one wave actually fared better 
in the next.

```{r hypothesis_3, fig.cap="H3: Did the same counties bear disproportionate burden across waves?"}

# Pivot to wide format: one row per county, one column per wave
wave_ranks <- wave_county_rates %>%
  dplyr::select(fips, county_name, state_name,
                pandemic_wave, wave_death_rate) %>%
  pivot_wider(
    names_from  = pandemic_wave,
    values_from = wave_death_rate
  ) %>%
  drop_na()

# Compute Spearman correlations between all wave pairs
wave_cols <- c("1_initial", "2_summer2020", "3_winter2021",
               "4_alpha", "5_delta", "6_omicron")

wave_cor <- wave_ranks %>%
  dplyr::select(all_of(wave_cols)) %>%
  cor(method = "spearman", use = "complete.obs")

# Display as readable table
wave_cor %>%
  as.data.frame() %>%
  rownames_to_column("Wave") %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = paste(
    "H3: Spearman Rank Correlations of County Death Rates Across Waves",
    "(1.0 = identical ranking, 0 = no relationship, negative = reversed)"
  )) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### H3 visual companion to the correlation table (focus on Wave 1 vs Delta)

```{r hypothesis_3_visual, fig.cap="H3: Wave 1 vs Delta burden -- did high-burden counties persist?"}

wave_ranks %>%
  filter(!is.na(`1_initial`), !is.na(`5_delta`)) %>%
  mutate(
    rank_wave1 = rank(`1_initial`),
    rank_delta = rank(`5_delta`),
    rank_diff  = rank_delta - rank_wave1,
    burden_persistence = case_when(
      rank_wave1 > quantile(rank_wave1, 0.75) &
        rank_delta > quantile(rank_delta, 0.75) ~ "Persistently High",
      rank_wave1 < quantile(rank_wave1, 0.25) &
        rank_delta < quantile(rank_delta, 0.25) ~ "Persistently Low",
      rank_wave1 > quantile(rank_wave1, 0.75) &
        rank_delta < quantile(rank_delta, 0.25) ~ "High->Low (improved)",
      rank_wave1 < quantile(rank_wave1, 0.25) &
        rank_delta > quantile(rank_delta, 0.75) ~ "Low->High (worsened)",
      TRUE                                       ~ "Middle"
    )
  ) %>%
  ggplot(aes(x = rank_wave1, y = rank_delta, color = burden_persistence)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  scale_color_manual(
    values = c(
      "Persistently High"   = "#E63946",
      "Persistently Low"    = "#2A9D8F",
      "High->Low (improved)"= "#457B9D",
      "Low->High (worsened)"= "#F4A261",
      "Middle"              = "grey70"
    ),
    name = NULL
  ) +
  labs(
    title    = "H3: County Burden Persistence -- Wave 1 vs Delta",
    subtitle = "Persistently high-burden counties appear in red upper right",
    x        = "County Death Rate Rank in Wave 1",
    y        = "County Death Rate Rank in Delta Wave",
    caption  = "Dashed line = identical rank in both waves"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Reading the H3 Scatter Plot
```
The near-zero Spearman correlation between Wave 1 and Delta county 
death rate ranks (ρ = -0.003) is visible in the broad scatter of 
grey points around the diagonal reference line — if burden had 
persisted geographically the points would cluster tightly along 
that dashed line. Instead the cloud is diffuse, confirming that 
knowing a county's Wave 1 rank tells you almost nothing about 
its Delta rank.

Three analytically distinct county groups emerge from the quadrant 
classification:

**Persistently High (red, upper right):** Counties ranking in the 
top quartile for both waves. These represent communities with 
structural vulnerabilities — demographic, economic, or geographic — 
that made them targets for successive variant waves regardless of 
the pandemic's shifting geography. Identifying what these counties 
share is a priority for the regression model in Section 7.

**High to Low / Improved (blue, lower right):** Counties with 
high Wave 1 burden that achieved relative protection by Delta. 
As established in H1, these are predominantly the urban Northeast 
corridor and Deep South counties where catastrophic Wave 1 exposure 
created population-level immunity ahead of Delta's arrival.

**The vertical spread at low Wave 1 ranks (left edge):** Counties 
with near-zero Wave 1 mortality were distributed across the entire 
spectrum of Delta outcomes — some remained protected, others became 
Delta epicenters. This pattern is consistent with the geographic 
reorganization hypothesis: initial sparing from Wave 1 conferred 
no protection advantage and in some cases left populations 
immunologically naive heading into the more transmissible Delta wave.
```

```{r h3_persistently_high, include = FALSE}

# ============================================================
# Identify the Persistently High burden counties
# These are the communities structural vulnerability kept
# at elevated risk across multiple variant waves
# ============================================================

persistently_high <- wave_ranks %>%
  filter(!is.na(`1_initial`), !is.na(`5_delta`)) %>%
  mutate(
    rank_wave1 = rank(`1_initial`),
    rank_delta = rank(`5_delta`),
    n_counties = n()
  ) %>%
  filter(
    rank_wave1 > quantile(rank_wave1, 0.75),
    rank_delta > quantile(rank_delta, 0.75)
  ) %>%
  left_join(
    county_weekly %>%
      dplyr::select(fips, pct_black, pct_65_over,
                    poverty_rate, pct_no_cars,
                    median_age, total_pop, pct_less_than_hs) %>%
      distinct(fips, .keep_all = TRUE),
    by = "fips"
  ) %>%
  arrange(desc(`5_delta`))

cat("Persistently High burden counties:", nrow(persistently_high), "\n\n")

# State distribution
cat("State distribution of Persistently High counties:\n")
persistently_high %>%
  count(state_name, sort = TRUE) %>%
  slice_head(n = 15) %>%
  print()

# Demographic profile vs all counties
cat("\nDemographic profile: Persistently High vs All Counties\n")
cat("--------------------------------------------------------\n")
tibble(
  characteristic   = c("Median Age", "% Age 65+", "% Black",
                        "Poverty Rate", "% Without HS Diploma", "% No Car", "Population"),
  persistently_high = c(
    round(mean(persistently_high$median_age,       na.rm = TRUE), 1),
    round(mean(persistently_high$pct_65_over,      na.rm = TRUE), 3),
    round(mean(persistently_high$pct_black,        na.rm = TRUE), 3),
    round(mean(persistently_high$poverty_rate,     na.rm = TRUE), 3),
    round(mean(persistently_high$pct_less_than_hs, na.rm = TRUE), 3),
    round(mean(persistently_high$pct_no_cars,      na.rm = TRUE), 3),
    round(mean(persistently_high$total_pop,        na.rm = TRUE), 0)
  ),
  all_counties = c(
    round(mean(wave_ranks$median_age,       na.rm = TRUE), 1),
    round(mean(wave_ranks$pct_65_over,      na.rm = TRUE), 3),
    round(mean(wave_ranks$pct_black,        na.rm = TRUE), 3),
    round(mean(wave_ranks$poverty_rate,     na.rm = TRUE), 3),
    round(mean(wave_ranks$pct_less_than_hs, na.rm = TRUE), 3),
    round(mean(wave_ranks$pct_no_cars,      na.rm = TRUE), 3),
    round(mean(wave_ranks$total_pop,        na.rm = TRUE), 0)
  )
) %>%
  kable(caption = "Demographic Profile: Persistently High Burden Counties vs All Counties") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r h3_persistently_high_fixed, echo=FALSE, results='hide'}
# Pull demographic reference for ALL counties from county_weekly
all_county_demos <- county_weekly %>%
  dplyr::select(fips, median_age, pct_65_over, pct_black,
                poverty_rate, pct_no_cars, total_pop, pct_less_than_hs) %>%
  distinct(fips, .keep_all = TRUE)

persistently_high_demos <- persistently_high %>%
  dplyr::select(fips) %>%
  left_join(all_county_demos, by = "fips")

tibble(
  Characteristic    = c("Median Age", "% Age 65+", "% Black",
                         "Poverty Rate", "% Without HS Diploma",
                         "% No Car", "Population"),
  Persistently_High = c(
    round(mean(persistently_high_demos$median_age,    na.rm = TRUE), 1),
    scales::percent(mean(persistently_high_demos$pct_65_over,  na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$pct_black,    na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$poverty_rate, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$pct_less_than_hs, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$pct_no_cars,  na.rm = TRUE), accuracy = 0.1),
    scales::comma(round(mean(persistently_high_demos$total_pop, na.rm = TRUE), 0))
  ),
  All_Counties = c(
    round(mean(all_county_demos$median_age,    na.rm = TRUE), 1),
    scales::percent(mean(all_county_demos$pct_65_over,  na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$pct_black,    na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$poverty_rate, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$pct_less_than_hs, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$pct_no_cars,  na.rm = TRUE), accuracy = 0.1),
    scales::comma(round(mean(all_county_demos$total_pop, na.rm = TRUE), 0))
  )
) %>%
  kable(caption = "Demographic Profile: Persistently High Burden vs All Counties") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r h3_burden_shift_comparison}
# ============================================================
# Three-way demographic comparison:
# Persistently High vs High-to-Low vs All Counties
# Tests whether High-to-Low counties differ structurally
# from both persistently burdened and average communities
# ============================================================

# Identify High-to-Low counties from wave_ranks
high_to_low <- wave_ranks %>%
  filter(!is.na(`1_initial`), !is.na(`5_delta`)) %>%
  mutate(
    rank_wave1 = rank(`1_initial`),
    rank_delta = rank(`5_delta`)
  ) %>%
  filter(
    rank_wave1 > quantile(rank_wave1, 0.75),  # top quartile Wave 1
    rank_delta < quantile(rank_delta, 0.25)    # bottom quartile Delta
  )

cat("High-to-Low counties identified:", nrow(high_to_low), "\n")

# Get demographics for High-to-Low counties
high_to_low_demos <- high_to_low %>%
  dplyr::select(fips) %>%
  left_join(all_county_demos, by = "fips")

# State distribution of High-to-Low counties
cat("\nState distribution of High-to-Low counties:\n")
high_to_low %>%
  count(state_name, sort = TRUE) %>%
  slice_head(n = 10) %>%
  print()

# Three-way comparison table
tibble(
  Characteristic = c(
    "Median Age",
    "% Age 65+",
    "% Black",
    "Poverty Rate",
    "% Without HS Diploma",
    "% No Car",
    "Population"
  ),
  Persistently_High = c(
    round(mean(persistently_high_demos$median_age,       na.rm = TRUE), 1),
    scales::percent(mean(persistently_high_demos$pct_65_over,      na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$pct_black,        na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$poverty_rate,     na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$pct_less_than_hs, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(persistently_high_demos$pct_no_cars,      na.rm = TRUE), accuracy = 0.1),
    scales::comma(round(mean(persistently_high_demos$total_pop,    na.rm = TRUE), 0))
  ),
  High_to_Low = c(
    round(mean(high_to_low_demos$median_age,       na.rm = TRUE), 1),
    scales::percent(mean(high_to_low_demos$pct_65_over,      na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(high_to_low_demos$pct_black,        na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(high_to_low_demos$poverty_rate,     na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(high_to_low_demos$pct_less_than_hs, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(high_to_low_demos$pct_no_cars,      na.rm = TRUE), accuracy = 0.1),
    scales::comma(round(mean(high_to_low_demos$total_pop,    na.rm = TRUE), 0))
  ),
  All_Counties = c(
    round(mean(all_county_demos$median_age,       na.rm = TRUE), 1),
    scales::percent(mean(all_county_demos$pct_65_over,      na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$pct_black,        na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$poverty_rate,     na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$pct_less_than_hs, na.rm = TRUE), accuracy = 0.1),
    scales::percent(mean(all_county_demos$pct_no_cars,      na.rm = TRUE), accuracy = 0.1),
    scales::comma(round(mean(all_county_demos$total_pop,    na.rm = TRUE), 0))
  )
) %>%
  kable(caption = paste0(
    "Three-Way Demographic Comparison: ",
    "Persistently High Burden vs High-to-Low vs All Counties"
  )) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(2, background = "#ffe0e0") %>%  # red tint for persistently high
  column_spec(3, background = "#e0f0ff") %>%  # blue tint for high to low
  column_spec(4, background = "#f5f5f5")       # grey for all counties
```

### Reading the Three-Way Comparison

The demographic profile of High-to-Low counties reveals 
something the scatter plot could only show visually — the 
communities that achieved the greatest protection during 
Delta were not structurally similar to the communities that 
remained persistently burdened. They were structurally 
different in almost every meaningful way.

High-to-Low counties had nearly ten times the population 
of Persistently High counties (392,205 vs 42,577), 
substantially lower poverty rates (10.8% vs 19.5% — and 
notably below the national average of 15.3%), and the 
lowest educational disadvantage of any group (6.9% without 
a high school diploma vs 8.9% nationally and 12.2% in 
Persistently High counties). Their elderly population share 
was actually below the national average at 16.5%.

These are not communities that achieved protection despite 
structural disadvantage. These are communities that entered 
the Delta wave with structural advantages — larger 
populations, better-resourced public health infrastructure, 
lower poverty, higher educational attainment, and the 
accumulated prior exposure of dense urban Wave 1 outbreaks. 
All of those factors pointed in the same protective 
direction simultaneously.

The Persistently High counties had none of those advantages. 
Small populations, high poverty, lower educational 
attainment, majority minority demographics, and rural 
geographic isolation combined to create communities that 
neither prior exposure nor vaccination nor healthcare 
infrastructure could adequately protect across multiple 
successive waves.

> **The Central Finding of This Comparison:** The pandemic 
> did not treat structurally disadvantaged communities 
> differently because of random chance or bad luck. It 
> treated them differently because the conditions that make 
> a community resilient during a health emergency — economic 
> resources, educational investment, healthcare 
> infrastructure, population density sufficient to generate 
> prior immunity — were systematically absent in the 
> communities that bore persistent burden. Those conditions 
> were not created by COVID. They existed before the first 
> case was ever recorded. COVID simply made their 
> consequences impossible to ignore.
>
> The question this analysis raises — and that Phase 2 data 
> integration will pursue — is not why these communities 
> fared worse during COVID. The data has answered that. The 
> question is what investment in the structural conditions 
> identified here would cost relative to the human and 
> economic price of the next pandemic finding them in the 
> same state.

### Reading the H3 Results

The correlation matrix delivers one of the most striking findings 
in the entire analysis. Wave 1 is essentially uncorrelated with 
every subsequent wave except Summer 2020:

- Wave 1 vs Delta: ρ = -0.003 (effectively zero)
- Wave 1 vs Omicron: ρ = 0.013 (effectively zero)
- Wave 1 vs Winter 2021: ρ = -0.006 (effectively zero)

Knowing which counties suffered most in the spring of 2020 
tells you almost nothing about which counties suffered most 
in the Delta wave eighteen months later. The pandemic 
completely reorganized geographically after its initial 
outbreak — the communities that bore the first wave's burden 
were largely spared subsequent waves while communities that 
were relatively protected in 2020 became epicenters of Delta 
and beyond.

The later waves — Winter 2021, Alpha, Delta, and Omicron — 
show moderate positive correlations with each other ranging 
from 0.242 to 0.389. A stable geography of vulnerability 
emerged from Winter 2021 onward and persisted through the 
end of the analysis period. Once a community entered the 
high-burden tier in the second year of the pandemic it tended 
to stay there.

**Delta was the pivot point.** It correlates near zero with 
Wave 1 (ρ = -0.003) but moderately with Omicron 
(ρ = 0.389) — meaning Delta completed the geographic 
reorganization that Summer 2020 had begun and established 
the burden geography that defined the remainder of the 
pandemic.

The burden persistence scatter plot makes this visible at 
the county level. The broad scatter of grey points around 
the diagonal reference line — rather than clustering along 
it — is the visual proof of near-zero correlation. Counties 
that improved dramatically from Wave 1 to Delta appear in 
blue in the lower right — these are your Urban Northeast 
and Deep South protected clusters from H1. Counties that 
remained consistently burdened across both waves appear in 
red in the upper right — these are the structurally 
vulnerable communities whose characteristics the regression 
model in Section 7 formally identifies.

> **The Three Hypotheses Together:** H1 showed that counties 
> hardest hit in Wave 1 achieved substantial protection by 
> Delta through prior immunity mechanisms. H3 confirms the 
> statistical signature of that protection — near-zero 
> correlation between Wave 1 and Delta county rankings means 
> the burden did not just shift at the margins, it completely 
> reorganized. H2 explains why this reorganization had such 
> severe consequences — the case-to-death relationship 
> remained steep during Delta before the post-Omicron 
> structural decline. When burden shifted to counties with 
> lower prior exposure, lower vaccination rates, and less 
> healthcare infrastructure, each case was still highly 
> lethal. The three findings interlock into a single coherent 
> story about how structural disadvantage determined which 
> communities paid the price for each successive wave.

# 4. Excess Mortality -- Counties Above Expected Death Rate

```{r county_totals}
county_totals <- county_weekly %>%
  group_by(fips, county_name, state_name,
           total_pop, median_age, pct_white, pct_black, pct_hispanic,
           pct_asian, pct_65_over, poverty_rate,
           pct_no_cars, pop_per_housing_unit, pct_less_than_hs,
           pct_bachelors_plus, pct_ag_mining_workers) %>%
  summarise(
    total_confirmed = max(cumulative_confirmed, na.rm = TRUE),
    total_deceased  = max(cumulative_deceased,  na.rm = TRUE),
    cum_case_rate   = max(cum_case_rate_100k,   na.rm = TRUE),
    cum_death_rate  = max(cum_death_rate_100k,  na.rm = TRUE),
    cfr_overall     = max(cumulative_deceased, na.rm = TRUE) /
                      pmax(max(cumulative_confirmed, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  filter(total_pop >= 5000)
```

```{r excess_mortality}
national_mean_dr <- mean(county_totals$cum_death_rate, na.rm = TRUE)
national_sd_dr   <- sd(county_totals$cum_death_rate,   na.rm = TRUE)

county_totals <- county_totals %>%
  mutate(
    death_rate_zscore = (cum_death_rate - national_mean_dr) / national_sd_dr,
    excess_flag = case_when(
      death_rate_zscore >= 3  ~ "Extreme (>=3 SD)",
      death_rate_zscore >= 2  ~ "High (2-3 SD)",
      death_rate_zscore >= 1  ~ "Elevated (1-2 SD)",
      death_rate_zscore >= -1 ~ "Average",
      TRUE                    ~ "Below Average"
    ),
    excess_flag = factor(excess_flag, levels = c(
      "Extreme (>=3 SD)", "High (2-3 SD)", "Elevated (1-2 SD)",
      "Average", "Below Average"
    ))
  )

county_totals %>%
  count(excess_flag) %>%
  mutate(pct = scales::percent(n / sum(n), accuracy = 0.1)) %>%
  kable(caption = "County Distribution by Excess Mortality Flag") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r excess_top_counties}
county_totals %>%
  arrange(desc(cum_death_rate)) %>%
  slice_head(n = 30) %>%
  dplyr::select(county_name, state_name, total_pop, cum_death_rate,
         death_rate_zscore, excess_flag, pct_65_over, poverty_rate) %>%
  mutate(
    cum_death_rate    = round(cum_death_rate, 1),
    death_rate_zscore = round(death_rate_zscore, 2),
    pct_65_over       = scales::percent(pct_65_over, accuracy = 0.1),
    poverty_rate      = scales::percent(poverty_rate, accuracy = 0.1)
  ) %>%
  kable(caption = "Top 30 Counties by Cumulative COVID Death Rate per 100k") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Persistence Overlap

How much do excess mortality flags overlap with persistently high burden counties from H3? This connects Section 4 findings to H3 findings formally.

```{r excess_persistence_overlap}

overlap_check <- county_totals %>%
  dplyr::select(fips, county_name, state_name,
                excess_flag, cum_death_rate) %>%
  left_join(
    persistently_high %>%
      dplyr::select(fips) %>%
      mutate(persistently_high = TRUE),
    by = "fips"
  ) %>%
  mutate(persistently_high = replace_na(persistently_high, FALSE))

# Cross-tabulation
overlap_check %>%
  group_by(excess_flag, persistently_high) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(
    names_from  = persistently_high,
    values_from = n,
    names_prefix = "persist_high_"
  ) %>%
  mutate(
    persist_high_FALSE = replace_na(persist_high_FALSE, 0),
    persist_high_TRUE  = replace_na(persist_high_TRUE,  0),
    total              = persist_high_FALSE + persist_high_TRUE,
    pct_also_persistent = scales::percent(
      persist_high_TRUE / total, accuracy = 0.1)
  ) %>%
  rename(
    Excess_Flag          = excess_flag,
    Not_Persistent       = persist_high_FALSE,
    Also_Persistent_High = persist_high_TRUE,
    Total                = total,
    Pct_Also_Persistent  = pct_also_persistent
  ) %>%
  kable(caption = "Overlap: Excess Mortality Flags and Persistently High Burden Counties") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
### Key Finding

> Monotonic relationship between excess mortality severity and wave
> persistence confirms these are not measurement artifacts. The same
> structural vulnerabilities that produced extreme cumulative death rates
> also prevented recovery between waves. The 0% overlap at Below Average
> provides a clean lower boundary validating both classifications.


# 4b. Persistently High -- Counties Above Expected Death Rate

> **Identify which states have complete, partial, or zero county coverage
> in the dataset** 

```{r missing_states_diagnosis}

# Get all US counties from tigris shapefile
# all_us_counties pre-loaded in tigris_setup chunk
# No download needed -- using cached version

# Compare against county_totals
coverage_by_state <- all_us_counties %>%
  left_join(
    county_totals %>%
      dplyr::select(fips, cum_death_rate) %>%
      mutate(in_dataset = TRUE),
    by = "fips"
  ) %>%
  mutate(in_dataset = replace_na(in_dataset, FALSE)) %>%
  group_by(STATE_NAME) %>%
  summarise(
    total_counties    = n(),
    counties_in_data  = sum(in_dataset),
    counties_missing  = sum(!in_dataset),
    pct_coverage      = scales::percent(
      sum(in_dataset) / n(), accuracy = 0.1),
    .groups = "drop"
  ) %>%
  arrange(counties_missing %>% desc())

# States with zero coverage
cat("States with ZERO county coverage:\n")
coverage_by_state %>%
  filter(counties_in_data == 0) %>%
  print()

cat("\nStates with PARTIAL coverage:\n")
coverage_by_state %>%
  filter(counties_in_data > 0, counties_missing > 0) %>%
  print()

cat("\nStates with COMPLETE coverage:\n")
coverage_by_state %>%
  filter(counties_missing == 0) %>%
  print()
```

> **Potential Coverage Bias:** Assess whether missing states/counties bias
> findings directionally. Compare demographic profile of missing versus
> included counties (document potential limitations).

```{r coverage_bias_assessment}

# Compare demographic profile of missing vs present counties
missing_fips <- all_us_counties %>%
  left_join(
    county_totals %>%
      dplyr::select(fips) %>%
      mutate(in_dataset = TRUE),
    by = "fips"
  ) %>%
  mutate(in_dataset = replace_na(in_dataset, FALSE)) %>%
  filter(!in_dataset) %>%
  pull(fips)

# Get ACS demographics for missing counties
missing_demos <- county_weekly %>%
  dplyr::select(fips, total_pop, pct_black, pct_65_over,
                poverty_rate, median_age, pct_no_cars, pct_less_than_hs) %>%
  distinct(fips, .keep_all = TRUE) %>%
  mutate(coverage = if_else(fips %in% missing_fips,
                            "Missing from Analysis",
                            "Included in Analysis"))

missing_demos %>%
  group_by(coverage) %>%
  summarise(
    n_counties    = n(),
    median_pop    = scales::comma(round(median(total_pop,    na.rm = TRUE), 0)),
    mean_pct_black = scales::percent(mean(pct_black,   na.rm = TRUE), accuracy = 0.1),
    mean_poverty  = scales::percent(mean(poverty_rate, na.rm = TRUE), accuracy = 0.1),
    mean_pct_65_over  = scales::percent(mean(pct_65_over,  na.rm = TRUE), accuracy = 0.1),
    mean_pct_less_than_hs = scales::percent(mean(pct_less_than_hs, na.rm = TRUE), accuracy = 0.1),
    .groups = "drop"
  ) %>%
  kable(caption = paste(
    "Demographic Comparison: Counties Included vs Missing from Analysis",
    "-- Directional bias assessment"
  )) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Note: our county_totals already filters total_pop >= 5000
# which excludes the most unstable small-county rate estimates
# The missing counties with median pop 2,860 would largely
# be excluded by this filter anyway -- further reducing
# the practical impact of the coverage gap on core findings
```

**Interactive Persistently High Burden Counties Map**

```{r leaflet_persistently_high_map}
library(leaflet)
library(sf)

# Get county shapefile for just the persistently high counties
options(tigris_use_cache = TRUE)
ph_sf <- counties_sf %>%
  filter(fips %in% persistently_high$fips) %>%
  left_join(
    persistently_high %>%
      dplyr::select(fips, county_name, state_name,
                    `1_initial`, `5_delta`,
                    pct_black, pct_65_over,
                    poverty_rate, total_pop) %>%
      left_join(
        all_county_demos %>%
          dplyr::select(fips, pct_less_than_hs),
        by = "fips"
      ),
    by = "fips"
  )

# Color palette based on Wave 1 death rate
pal <- colorNumeric(
  palette = "YlOrRd",
  domain  = ph_sf$`1_initial`,
  na.color = "grey80"
)

# Build tooltip
ph_sf <- ph_sf %>%
  mutate(
    popup_text = paste0(
      "<b>", county_name, ", ", state_name, "</b><br>",
      "<b style='color:#E63946'>Persistently High Burden County</b><br>",
      "────────────────────<br>",
      "Wave 1 Death Rate: ", round(`1_initial`, 1), " per 100k<br>",
      "Delta Death Rate: ",  round(`5_delta`,   1), " per 100k<br>",
      "────────────────────<br>",
      "Population: ", scales::comma(total_pop), "<br>",
      "% Black: ", scales::percent(pct_black, accuracy = 0.1), "<br>",
      "% Age 65+: ", scales::percent(pct_65_over, accuracy = 0.1), "<br>",
      "Poverty Rate: ", scales::percent(poverty_rate, accuracy = 0.1), "<br>",
      "% Without HS Diploma: ",
      scales::percent(pct_less_than_hs, accuracy = 0.1)
    )
  )

leaflet(ph_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor   = ~pal(`1_initial`),
    fillOpacity = 0.8,
    color       = "white",
    weight      = 1.5,
    popup       = ~popup_text,
    highlight   = highlightOptions(
      weight      = 3,
      color       = "#E63946",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal      = pal,
    values   = ~`1_initial`,
    title    = "Wave 1 Death Rate<br>per 100k",
    position = "bottomright"
  ) %>%
  addControl(
    html = paste0(
      "<div style='background:white; padding:8px; ",
      "border-radius:4px; font-size:12px;'>",
      "<b>157 Persistently High Burden Counties</b><br>",
      "Top quartile death rate in BOTH<br>",
      "Wave 1 and Delta wave<br>",
      "<i>Click any county for demographic profile</i>",
      "</div>"
    ),
    position = "topleft"
  )
```

**Choropleth: Cumulative COVID Death Rate per 100K by County**

```{r national_excess_map, fig.cap="Choropleth: Cumulative COVID Death Rate per 100k by County"}
invisible(gc())  # trigger garbage collection silently -- suppresses memory table output
# counties_sf pre-loaded in tigris_setup chunk
# No download needed -- using cached version

map_data <- counties_sf %>%
  left_join(county_totals, by = "fips")

ggplot(map_data) +
  geom_sf(aes(fill = cum_death_rate), color = NA) +
  scale_fill_viridis_c(
    option    = "magma",
    direction = -1,
    name      = "Deaths\nper 100k",
    na.value  = "grey85",
    labels    = comma
  ) +
  labs(
    title    = "Cumulative COVID-19 Death Rate per 100k by County",
    subtitle = "March 2020 - September 2022 | Grey = no data in source pipeline",
    caption  = paste0(
      "Source: Google COVID-19 Open Data + ACS 2019 5-Year | ",
    "Coverage: 2,904 of ~3,143 US counties (92%) | ",
    "6 states have zero coverage (AL, AR, AZ, CA, CO, CT) -- ",
    "see Section 11 for bias assessment | ",
    "Interactive state-level map planned for Phase 2 ",
    "with complete national coverage"
    )
  ) +
  theme_void(base_size = 13) +
  theme(legend.position = "right")
```


# 5. Case-Fatality Relationship by County

> **Relationship Between Cumulative Case Rate and Death Rate by County**

```{r cfr_scatter, fig.cap="Relationship between cumulative case rate and death rate by county"}
library(plotly)

# Build tooltip text for county points
cfr_plot_data <- county_totals %>%
  filter(!is.na(cfr_overall), cfr_overall < 0.3) %>%
  mutate(
    tooltip_text = paste0(
      "<b>", county_name, ", ", state_name, "</b><br>",
      "Death Rate: ",  round(cum_death_rate, 1), " per 100k<br>",
      "Case Rate: ",   round(cum_case_rate,  1), " per 100k<br>",
      "CFR: ",         scales::percent(cfr_overall, accuracy = 0.1), "<br>",
      "Population: ",  scales::comma(total_pop), "<br>",
      "% Age 65+: ",   scales::percent(pct_65_over,  accuracy = 0.1), "<br>",
      "Poverty Rate: ",scales::percent(poverty_rate, accuracy = 0.1), "<br>",
      "% Black: ",     scales::percent(pct_black,    accuracy = 0.1)
    )
  )

# Fit the regression model explicitly to build
# tooltip-bearing reference points along the line
lm_fit    <- lm(cum_death_rate ~ cum_case_rate, data = cfr_plot_data)
lm_r2     <- summary(lm_fit)$r.squared
lm_slope  <- coef(lm_fit)[2]
lm_int    <- coef(lm_fit)[1]
lm_pval   <- summary(lm_fit)$coefficients[2, 4]

# Build 50 reference points along the regression line
# These carry the tooltip and are rendered as invisible points
case_range <- seq(min(cfr_plot_data$cum_case_rate, na.rm = TRUE),
                  max(cfr_plot_data$cum_case_rate, na.rm = TRUE),
                  length.out = 50)

lm_line_data <- tibble(
  cum_case_rate  = case_range,
  cum_death_rate = lm_int + lm_slope * case_range,
  tooltip_text   = paste0(
    "<b>National OLS Regression Line</b><br>",
    "At this case rate: ",
    scales::comma(round(case_range, 0)), " per 100k<br>",
    "Expected death rate: ",
    round(lm_int + lm_slope * case_range, 1), " per 100k<br>",
    "────────────────────<br>",
    "Slope: ", round(lm_slope, 4),
    " deaths per case (per 100k)<br>",
    "Intercept: ", round(lm_int, 1), "<br>",
    "R²: ", round(lm_r2, 3), "<br>",
    "p-value: ", scales::pvalue(lm_pval), "<br>",
    "────────────────────<br>",
    "<i>Points above this line = higher than",
    " expected death rate</i><br>",
    "<i>Points below = lower than expected</i>"
  )
)

# Build confidence interval band data
pred_data  <- predict(lm_fit,
                      newdata   = tibble(cum_case_rate = case_range),
                      interval  = "confidence",
                      level     = 0.95) %>%
  as_tibble() %>%
  mutate(
    cum_case_rate = case_range,
    tooltip_text  = paste0(
      "<b>95% Confidence Band</b><br>",
      "At case rate: ",
      scales::comma(round(case_range, 0)), " per 100k<br>",
      "Expected death rate: ", round(fit,  1), " per 100k<br>",
      "95% CI lower: ",        round(lwr,  1), " per 100k<br>",
      "95% CI upper: ",        round(upr,  1), " per 100k<br>",
      "────────────────────<br>",
      "<i>Band width reflects uncertainty in the",
      " national average at each case rate level</i>"
    )
  )

# Build the plot
p_cfr <- ggplot() +

  # Confidence band as ribbon -- invisible fill, tooltip via points
  geom_ribbon(data = pred_data,
              aes(x = cum_case_rate, ymin = lwr, ymax = upr),
              fill = "grey70", alpha = 0.25, inherit.aes = FALSE) +

  # Invisible points along CI band upper edge for tooltips
  geom_point(data = pred_data,
             aes(x = cum_case_rate, y = upr, text = tooltip_text),
             size = 0, alpha = 0, inherit.aes = FALSE) +

  # Invisible points along CI band lower edge for tooltips
  geom_point(data = pred_data,
             aes(x = cum_case_rate, y = lwr, text = tooltip_text),
             size = 0, alpha = 0, inherit.aes = FALSE) +

  # County scatter points
  geom_point(data = cfr_plot_data,
             aes(x     = cum_case_rate,
                 y     = cum_death_rate,
                 color = pct_65_over,
                 size  = total_pop,
                 text  = tooltip_text),
             alpha = 0.6) +

  # Regression line as visible dashed line
  geom_line(data = lm_line_data,
            aes(x = cum_case_rate, y = cum_death_rate),
            color    = "black",
            linewidth = 0.8,
            linetype = "dashed",
            inherit.aes = FALSE) +

  # Invisible points along regression line carrying tooltip
  geom_point(data = lm_line_data,
             aes(x = cum_case_rate, y = cum_death_rate,
                 text = tooltip_text),
             size = 0, alpha = 0, inherit.aes = FALSE) +

  scale_color_viridis_c(
    name   = "% Age 65+",
    labels = percent,
    option = "viridis"
  ) +
  scale_size_continuous(
    name   = "Population",
    labels = comma,
    range  = c(0.5, 6)
  ) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Cumulative Case Rate vs. Death Rate by County",
    subtitle = paste0(
      "Color = % population 65+  |  Size = county population  |  ",
      "Hover points OR regression line for detail"
    ),
    x       = "Cumulative Cases per 100k",
    y       = "Cumulative Deaths per 100k",
    caption = "Dashed line = OLS fit | Grey band = 95% confidence interval"
  ) +
  theme_minimal(base_size = 13)

# Convert to interactive plotly
ggplotly(p_cfr, tooltip = "text") %>%
  layout(
    hoverlabel = list(
      bgcolor     = "white",
      font        = list(size = 12),
      bordercolor = "grey40"
    ),
    legend = list(orientation = "v", x = 1.05)
  )
```

```{r cfr_by_wave, fig.cap="Case fatality ratio over time by vaccine era"}
county_weekly %>%
  filter(new_deceased_wk >= 0, new_confirmed_wk > 0) %>%
  mutate(
    vaccine_era_label = recode(vaccine_era,
      "pre_vaccine"        = "Pre-Vaccine",
      "early_rollout"      = "Early Rollout",
      "broad_availability" = "Broad Availability"
    )
  ) %>%
  group_by(week_start, vaccine_era_label) %>%
  summarise(
    cfr_wk = sum(new_deceased_wk, na.rm = TRUE) /
              pmax(sum(new_confirmed_wk, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = week_start, y = cfr_wk,
             color = vaccine_era_label,
             group = vaccine_era_label)) +
  geom_line(alpha = 0.2, linewidth = 0.4) +
  geom_smooth(method    = "loess",
              se        = FALSE,
              linewidth = 1.8,
              span      = 0.3) +
  scale_color_manual(
    values = c(
      "Pre-Vaccine"        = "#E63946",
      "Early Rollout"      = "#457B9D",
      "Broad Availability" = "#2A9D8F"
    ),
    name = "Vaccine Era"
  ) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = percent, limits = c(0, 0.15)) +
  annotate("text",
           x     = as.Date("2020-04-15"),
           y     = 0.105,
           label = "Testing scarcity\nmasks true IFR",
           size  = 3.2,
           color = "#E63946",
           hjust = 0) +
  annotate("text",
           x     = as.Date("2021-02-01"),
           y     = 0.055,
           label = "Early doses\nto highest risk",
           size  = 3.2,
           color = "#457B9D",
           hjust = 0) +
  annotate("text",
           x     = as.Date("2022-01-01"),
           y     = 0.055,
           label = "Omicron\nattenuation",
           size  = 3.2,
           color = "#2A9D8F",
           hjust = 0) +
  labs(
    title    = "National Case Fatality Ratio Over Time",
    subtitle = "Smoothed weekly CFR by vaccine era",
    x        = NULL,
    y        = "Case Fatality Ratio"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```



# 6. Demographic Drivers of Death Rate

Before building the formal regression model in Section 7, this 
section examines how individual demographic characteristics relate 
to county COVID death rates. This serves as an analytical screening 
step — identifying which variables deserve inclusion in the model 
and previewing the direction of each relationship.

### Reading the Correlation Table

Pearson correlation measures the linear relationship between two 
variables on a scale from -1.0 to 1.0. A value close to 1.0 means 
the two variables tend to rise and fall together — as one increases 
the other increases. A value close to -1.0 means they move in 
opposite directions — as one increases the other decreases. A value 
near zero means little to no linear relationship exists between them.

The table below ranks each demographic variable by the strength of 
its relationship with cumulative COVID death rate. Variables near 
the top of the list are the strongest candidates for the regression 
model. Variables near the bottom have weaker relationships and may 
lose statistical significance after other factors are controlled for.

> **Important Limitation:** Correlation measures association, not 
> causation. A strong correlation between two variables does not 
> mean one caused the other. Many of these demographic 
> characteristics are themselves correlated with each other — 
> counties with high poverty rates also tend to have lower 
> educational attainment, for example. The regression model in 
> Section 7 formally separates these overlapping relationships 
> to identify which factors have independent predictive power 
> after controlling for the others.

```{r demographic_correlations}
demo_vars <- county_totals %>%
  dplyr::select(cum_death_rate, pct_65_over, pct_black, pct_hispanic,
         pct_white, poverty_rate, pct_no_cars, median_age,
         pop_per_housing_unit, pct_less_than_hs) %>%
  drop_na()

cor(demo_vars, use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  select(variable, cum_death_rate) %>%
  filter(variable != "cum_death_rate") %>%
  arrange(desc(abs(cum_death_rate))) %>%
  mutate(cum_death_rate = round(cum_death_rate, 3)) %>%
  rename(correlation_with_death_rate = cum_death_rate) %>%
  mutate(variable = recode(variable,
    "pct_less_than_hs"     = "% Without HS Diploma",
    "poverty_rate"         = "Poverty Rate",
    "pct_65_over"          = "% Age 65+",
    "pct_black"            = "% Black Population",
    "pct_no_cars"          = "% No Car Access",
    "pct_white"            = "% White Population",
    "median_age"           = "Median Age",
    "pop_per_housing_unit" = "Pop. per Housing Unit",
    "pct_hispanic"         = "% Hispanic Population"
  )) %>%
  kable(caption = "Pearson Correlation: Demographic Variables vs. Cumulative Death Rate") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### COVID Death Rates by Black Population Share

The correlation table identifies Black population share as the 
fourth strongest predictor of county COVID death rates. Combined 
with the finding that persistently high burden counties have nearly 
three times the national average Black population share (22.9% vs 
8.4%), a closer examination of how death rates vary across counties 
grouped by Black population concentration is warranted.

The chart below divides all counties into four equal groups — 
quartiles — ranked from lowest to highest Black population share. 
If Black population share has a meaningful relationship with COVID 
mortality the groups with higher Black population shares should 
show systematically higher death rates.

> **Analytical Note:** This analysis examines population-level 
> patterns at the county level — not individual-level outcomes. 
> A county with a high Black population share experiencing elevated 
> death rates does not mean Black residents specifically drove that 
> outcome. It means something about the structural characteristics 
> of counties with those demographic profiles — healthcare access, 
> poverty, educational attainment — predicted worse outcomes. The 
> regression model in Section 7 formally tests which structural 
> factors explain this pattern.

```{r race_boxplots, fig.cap="Death rate patterns by Black population quartile"}
library(patchwork)

# Prepare quartile data
race_data <- county_totals %>%
  mutate(
    pct_black_q = ntile(pct_black, 4),
    pct_black_q = factor(pct_black_q,
                         labels = c("Q1\n(0-3% Black)",
                                    "Q2\n(3-9% Black)",
                                    "Q3\n(9-25% Black)",
                                    "Q4\n(25-73% Black)"))
  ) %>%
  filter(!is.na(pct_black_q))

# Panel 1: Mean death rate with error bars -- clearest view of shift
p1 <- race_data %>%
  group_by(pct_black_q) %>%
  summarise(
    mean_dr = mean(cum_death_rate,  na.rm = TRUE),
    se_dr   = sd(cum_death_rate,    na.rm = TRUE) /
              sqrt(n()),
    n       = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = pct_black_q, y = mean_dr, fill = pct_black_q)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_dr - 1.96 * se_dr,
                    ymax = mean_dr + 1.96 * se_dr),
                width = 0.2, linewidth = 0.8) +
  geom_text(aes(label = paste0(round(mean_dr, 0), "\nper 100k")),
            vjust = -0.8, size = 3.2, fontface = "bold") +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(labels = comma,
                     limits = c(0, 500)) +
  labs(
    title    = "Mean Death Rate by Quartile",
    subtitle = "Error bars = 95% CI",
    x        = "% Black Population Quartile",
    y        = "Mean Deaths per 100k"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Panel 2: Proportion of extreme mortality counties per quartile
p2 <- race_data %>%
  mutate(extreme = cum_death_rate >
           (mean(cum_death_rate, na.rm = TRUE) +
            2 * sd(cum_death_rate, na.rm = TRUE))) %>%
  group_by(pct_black_q) %>%
  summarise(
    pct_extreme = mean(extreme, na.rm = TRUE),
    n           = n(),
    .groups     = "drop"
  ) %>%
  ggplot(aes(x = pct_black_q, y = pct_extreme, fill = pct_black_q)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = scales::percent(pct_extreme, accuracy = 0.1)),
            vjust = -0.5, size = 3.2, fontface = "bold") +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(labels = percent,
                     limits = c(0, 0.12)) +
  labs(
    title    = "% of Counties with Extreme Mortality",
    subtitle = ">2 SD above national mean",
    x        = "% Black Population Quartile",
    y        = "% of Counties Flagged Extreme"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Panel 3: Cumulative death rate density by quartile
p3 <- race_data %>%
  ggplot(aes(x = cum_death_rate,
             fill = pct_black_q,
             color = pct_black_q)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_brewer(palette  = "Reds", name = "Quartile") +
  scale_color_brewer(palette = "Reds", name = "Quartile") +
  scale_x_continuous(labels = comma, limits = c(0, 900)) +
  labs(
    title    = "Death Rate Distribution by Quartile",
    subtitle = "Right tail shift visible in Q3/Q4",
    x        = "Cumulative Deaths per 100k",
    y        = "Density"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

# Compose with patchwork
(p1 | p2) / p3 +
  plot_annotation(
    title   = "COVID-19 Death Rate by County Black Population Share",
    subtitle = paste0(
      "Disparity is concentrated in extreme mortality counties ",
      "rather than uniform shift across all counties"
    ),
    caption = "Quartile ranges: Q1=0-3%, Q2=3-9%, Q3=9-25%, Q4=25-73% Black population share"
  )
```



# 7. Excess Mortality Regression Model

To formally test whether structural factors predicted county COVID 
death rates after controlling for age, a Negative Binomial regression 
model was fitted to cumulative county death counts. This model was 
chosen over simpler alternatives because COVID death counts are 
overdispersed — the variation between counties is far greater than 
a standard Poisson model assumes. The Negative Binomial model 
handles this property correctly and produces honest confidence 
intervals as a result.

Counties with zero recorded deaths and populations below 5,000 
were excluded from the model. Very small counties produce 
statistically unstable rates — a single death in a community 
of 800 people generates a death rate that looks catastrophic 
per 100,000 but represents one family's loss rather than a 
systematic pattern.

> **How to Read the Results:** The model produces Incidence Rate 
> Ratios (IRR) for each variable. An IRR greater than 1.0 means 
> that factor is associated with a higher death rate. An IRR less 
> than 1.0 means it is associated with a lower death rate. An IRR 
> of exactly 1.0 means no relationship. The further a value sits 
> from 1.0 in either direction, the stronger the relationship.
>
> For example, an IRR of 3.7 for poverty rate means counties with 
> the highest poverty rates had approximately 3.7 times the expected 
> death rate of counties with the lowest poverty rates, after 
> accounting for every other variable in the model simultaneously.
>
> **Statistical significance** is indicated in the Significant 
> column. Variables marked Yes have relationships strong enough 
> that we can be confident they are real rather than the result 
> of random chance. Variables marked No — specifically Black 
> population share and car access — did not retain independent 
> statistical significance after structural factors were controlled 
> for. This is itself an important finding, explained below.


```{r nb_model}
model_data <- county_totals %>%
  dplyr::select(total_deceased, total_pop, pct_65_over, pct_black,
         pct_hispanic, poverty_rate, pct_no_cars, median_age,
         pop_per_housing_unit, pct_less_than_hs) %>%
  drop_na() %>%
  filter(total_deceased > 0, total_pop > 5000)

nb_model <- glm.nb(
  total_deceased ~
    offset(log(total_pop)) +
    pct_65_over +
    pct_black +
    pct_hispanic +
    poverty_rate +
    pct_no_cars +
    median_age +
    pop_per_housing_unit +
    pct_less_than_hs,
  data = model_data
)

summary(nb_model)
```

```{r model_irr_table}
tidy(nb_model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(across(c(estimate, conf.low, conf.high), ~round(., 3)),
         p.value     = scales::pvalue(p.value),
         significant = if_else(p.value < 0.05, "Yes", "No")) %>%
  rename(IRR = estimate, CI_low = conf.low, CI_high = conf.high) %>%
  kable(caption = "Negative Binomial Model: Incidence Rate Ratios") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### What the Model Found

The results challenge the dominant narrative of the pandemic 
response in important ways.

**Education was the strongest predictor.** Counties where more 
adults lack a high school diploma had dramatically higher COVID 
death rates — an IRR of 85.5 — after accounting for age, race, 
poverty, and every other variable simultaneously. This is not 
primarily a story about individual health decisions. Educational 
attainment at the county level reflects decades of investment 
in schools, economic opportunity, healthcare access, and 
community infrastructure. Counties with lower educational 
attainment had less of everything that protects a community 
during a health emergency.

**Age mattered, but less than education.** The federal COVID 
response was organized almost entirely around age as the primary 
risk factor — vaccine priority tiers, hospital protocols, and 
public messaging all centered on protecting elderly populations. 
Age was indeed a significant predictor with an IRR of 128. 
However its confidence interval is extremely wide (51 to 324) 
reflecting statistical uncertainty, while education's confidence 
interval is tighter (51 to 143) making it the more precisely 
estimated predictor of the two.

**Poverty independently predicted death rates.** An IRR of 3.7 
for poverty rate means high-poverty counties had nearly four 
times the expected death rate of low-poverty counties after 
controlling for everything else. Poverty and education are 
related but the model confirms they represent distinct 
dimensions of structural vulnerability — both matter 
independently.

**Hispanic population share was protective.** After controlling 
for poverty and education, counties with higher Hispanic 
population shares had lower expected death rates — an IRR of 
0.73. This is consistent with what epidemiologists call the 
Hispanic Health Paradox — Hispanic communities in the United 
States consistently show better health outcomes than their 
socioeconomic characteristics alone would predict.

**Black population share was not a statistically significant 
independent predictor.** This is the finding that most directly 
speaks to structural inequity. After controlling for educational 
attainment and poverty — the structural conditions that 
disproportionately characterize majority Black counties due to 
historical policy choices — Black population share itself had 
no statistically significant independent effect on death rates 
(IRR 0.958, p=0.425). The disparity documented throughout this 
analysis is real and severe. But it is not explained by race 
as an inherent risk factor. It is explained by the structural 
conditions that were imposed on these communities over 
generations and that COVID exposed with devastating clarity.

> **What This Means for Policy:** You cannot address the 
> mortality disparity documented in this analysis by targeting 
> race as a category. The data says the intervention points 
> are educational investment, poverty reduction, and healthcare 
> infrastructure in structurally disadvantaged communities — 
> which disproportionately but not exclusively serve majority 
> Black populations in the Deep South. Targeting the structural 
> conditions rather than the demographic category is both more 
> analytically defensible and more likely to produce lasting 
> change.

```{r irr_forest_plot, fig.cap="Forest plot of Incidence Rate Ratios"}
library(plotly)

irr_data <- tidy(nb_model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_clean = recode(term,
      pct_65_over          = "% Age 65+",
      pct_black            = "% Black Population",
      pct_hispanic         = "% Hispanic Population",
      poverty_rate         = "Poverty Rate",
      pct_no_cars          = "% No Car Access",
      median_age           = "Median Age",
      pop_per_housing_unit = "Pop. per Housing Unit",
      pct_less_than_hs     = "% Without HS Diploma"
    ),
    direction = if_else(estimate > 1, "Risk Factor", "Protective"),

    # Plain language interpretation for tooltip
    plain_language = case_when(
      term == "pct_less_than_hs" ~ paste0(
        "Counties with the highest share of adults\n",
        "without a high school diploma have\n",
        round(estimate, 1), "x the expected death rate\n",
        "of counties with the lowest share,\n",
        "after controlling for all other factors."
      ),
      term == "pct_65_over" ~ paste0(
        "Counties with the highest elderly population\n",
        "share have ", round(estimate, 1), "x the expected\n",
        "death rate of counties with the lowest share,\n",
        "after controlling for all other factors."
      ),
      term == "poverty_rate" ~ paste0(
        "Counties with the highest poverty rates\n",
        "have ", round(estimate, 2), "x the expected death rate\n",
        "of counties with the lowest poverty rates,\n",
        "after controlling for all other factors."
      ),
      term == "pop_per_housing_unit" ~ paste0(
        "Each additional person per housing unit\n",
        "multiplies the expected death rate by ",
        round(estimate, 3), ".\n",
        "Crowding independently amplifies transmission risk."
      ),
      term == "median_age" ~ paste0(
        "After controlling for % age 65+,\n",
        "higher median age is associated with\n",
        round((1 - estimate) * 100, 1), "% LOWER expected death rate.\n",
        "Likely reflects affluent retirement communities\n",
        "with better healthcare access."
      ),
      term == "pct_black" ~ paste0(
        "After controlling for education and poverty,\n",
        "Black population share has NO statistically\n",
        "significant independent effect (p=0.425).\n",
        "The disparity observed in Black communities\n",
        "is explained by structural conditions --\n",
        "not race as an independent risk factor."
      ),
      term == "pct_no_cars" ~ paste0(
        "No Car Access is NOT statistically significant\n",
        "after controlling for poverty and education.\n",
        "The wide confidence interval (0.32 to 2.26)\n",
        "reflects high uncertainty in this estimate."
      ),
      term == "pct_hispanic" ~ paste0(
        "After controlling for poverty and education,\n",
        "higher Hispanic population share is associated\n",
        "with ", round((1 - estimate) * 100, 1), "% LOWER expected death rate.\n",
        "Consistent with the Hispanic Health Paradox --\n",
        "better outcomes than socioeconomic factors predict."
      )
    ),

    # Full tooltip text
    tooltip_text = paste0(
      "<b>", term_clean, "</b><br>",
      "────────────────────────<br>",
      "<b>IRR: ", round(estimate, 3), "</b>",
      "  (", round(conf.low, 3), " to ", round(conf.high, 3), ")<br>",
      "p-value: ", scales::pvalue(p.value), "<br>",
      "Significant: ", if_else(p.value < 0.05, "Yes ✓", "No ✗"), "<br>",
      "────────────────────────<br>",
      "<i>", plain_language, "</i>"
    )
  )

# Build forest plot
p_forest <- ggplot(irr_data,
                   aes(x    = estimate,
                       y    = reorder(term_clean, estimate),
                       color = direction,
                       text  = tooltip_text)) +
  geom_vline(xintercept = 1,
             linetype   = "dashed",
             color      = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height    = 0.2,
                 color     = "grey40") +
  geom_point(size = 4) +
  scale_color_manual(
    values = c("Risk Factor" = "#E63946",
               "Protective"  = "#457B9D"),
    name   = NULL
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.5, 2, 5, 10, 50, 100, 200),
    labels = c("0.5", "0.75", "1", "1.5", "2", "5",
               "10", "50", "100", "200")
  ) +
  labs(
    title    = "Incidence Rate Ratios: County COVID-19 Death Rate",
    subtitle = "Negative Binomial Model | Log scale | Hover each point for plain language interpretation",
    x        = "Incidence Rate Ratio (IRR) — Log Scale",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

# Convert to interactive plotly
ggplotly(p_forest, tooltip = "text") %>%
  layout(
    hoverlabel = list(
      bgcolor     = "white",
      font        = list(size = 12),
      bordercolor = "grey40",
      align       = "left"
    ),
    legend = list(orientation = "h", y = -0.15)
  )
```

# 8. Seasonal Decomposition

Respiratory diseases have long followed seasonal patterns — 
winter brings people indoors, reduces ventilation, lowers 
vitamin D levels, and creates conditions that favor airborne 
transmission. Influenza, RSV, and the common cold all peak 
in the November through February window in the Northern 
Hemisphere. The question for COVID-19 is whether it followed 
the same pattern — and if so, what that tells us about the 
communities that were already structurally vulnerable when 
winter arrived.

To answer this, STL decomposition was applied to the national 
weekly death rate time series. STL — Seasonal-Trend 
decomposition using Loess — mathematically separates a time 
series into three distinct components:

- **Trend** — the underlying long-term direction of mortality, 
  smoothed across a 13-week window (one meteorological season) 
  to remove short-term noise
- **Seasonal** — the repeating annual cycle of higher winter 
  and lower summer mortality, extracted as the pattern that 
  recurs at the same time each year
- **Remainder** — what is left after trend and season are 
  removed; genuine anomalies that neither the long-term 
  direction nor the calendar can explain

STL was chosen over classical decomposition methods because 
it is robust to outliers — a pandemic with multiple extreme 
waves requires a decomposition method that can identify 
seasonal patterns without being distorted by the waves 
themselves.

> **Why This Matters for Structural Inequity:** If COVID 
> mortality followed consistent seasonal patterns, the 
> communities identified throughout this analysis as 
> persistently high burden were not just exposed to higher 
> baseline mortality — they were exposed to that higher 
> baseline amplified by the same winter seasonality as 
> everyone else. A community already at 400 deaths per 100k 
> cumulative burden entering November faces a different 
> winter than a community at 150 deaths per 100k. Seasonal 
> amplification compounds existing structural disadvantage.
````

```{r stl_decomposition, fig.cap="STL decomposition of national weekly death rate"}
# ============================================================
# SEASONAL PREP + STL DECOMPOSITION -- combined into one chunk
# to ensure stl_fit is available in the knit environment
# ============================================================

national_ts <- county_weekly %>%
  group_by(week_start) %>%
  summarise(
    death_rate_nat = weighted.mean(death_rate_wk,
                                   total_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(week_start) %>%
  filter(!is.na(death_rate_nat))

national_tsibble <- national_ts %>%
  mutate(week = yearweek(week_start)) %>%
  as_tsibble(index = week)

stl_fit <- national_tsibble %>%
  model(STL(death_rate_nat ~ trend(window = 13) +
              season(window = "periodic")))

# Extract components
stl_components <- components(stl_fit)

stl_df <- stl_components %>%
  as_tibble() %>%
  mutate(date = as.Date(week))

stl_long <- stl_df %>%
  dplyr::select(date, death_rate_nat, trend,
                season_year, remainder) %>%
  pivot_longer(
    cols      = c(death_rate_nat, trend,
                  season_year, remainder),
    names_to  = "component",
    values_to = "value"
  ) %>%
  mutate(
    component = factor(component,
      levels = c("death_rate_nat", "trend",
                 "season_year",    "remainder"),
      labels = c(
        "Observed Death Rate",
        "Long-Term Trend\n(13-week smoothing window)",
        "Seasonal Component\n(repeating winter/summer cycle)",
        "Remainder\n(unexplained after trend + season)"
      )
    )
  )

winter_bands <- tibble(
  xmin  = as.Date(c("2020-11-01", "2021-11-01")),
  xmax  = as.Date(c("2021-02-28", "2022-02-28")),
  label = c("Winter\n2020-21", "Winter\n2021-22")
)

ggplot(stl_long, aes(x = date, y = value)) +
  geom_rect(
    data        = winter_bands,
    aes(xmin = xmin, xmax = xmax,
        ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill        = "#AED6F1",
    alpha       = 0.25
  ) +
  geom_hline(yintercept = 0,
             linetype   = "dotted",
             color      = "grey60",
             linewidth  = 0.5) +
  geom_line(color = "#2C3E50", linewidth = 0.7) +
  geom_vline(
    data = data.frame(
      component = factor(
        "Remainder\n(unexplained after trend + season)",
        levels    = levels(stl_long$component)
      ),
      date = as.Date("2021-09-01")
    ),
    aes(xintercept = date),
    color       = "#E63946",
    linetype    = "dashed",
    linewidth   = 0.8,
    inherit.aes = FALSE
  ) +
  facet_wrap(~component,
             ncol           = 1,
             scales         = "free_y",
             strip.position = "left") +
  scale_x_date(
    date_breaks  = "3 months",
    date_labels  = "%b %Y",
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1)
  ) +
  labs(
    title    = "STL Decomposition: National Weekly COVID Death Rate",
    subtitle = paste0(
      "Blue shading = winter months (Nov-Feb) | ",
      "Red dashed line = Delta variant anomaly | ",
      "Dotted line = zero reference"
    ),
    x       = NULL,
    y       = "Deaths per 100k",
    caption = paste0(
      "STL = Seasonal-Trend decomposition via Loess | ",
      "Trend window = 13 weeks (one meteorological season) | ",
      "Seasonal window = periodic (same pattern assumed each year)"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.placement  = "outside",
    strip.text.y     = element_text(
      angle    = 0,
      hjust    = 1,
      size     = 9,
      face     = "bold",
      color    = "#2C3E50"
    ),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.spacing    = unit(0.8, "lines"),
    plot.subtitle    = element_text(size = 9, color = "grey40"),
    plot.caption     = element_text(size = 8, color = "grey50"),
    plot.margin      = margin(t = 10, r = 10,
                              b = 10, l = 120, unit = "pt")
  )
```

### Reading the Decomposition

The four panels work together to tell a complete story about 
what drove national COVID mortality at any given moment.

**Panel 1 — Observed Death Rate** is simply the raw weekly 
data — the actual measured mortality rate before any 
mathematical separation. The pandemic waves are all visible 
here: the initial spring 2020 outbreak, the summer 2020 
Sunbelt surge, the catastrophic winter 2020-2021 peak, the 
Delta spike, and the Omicron surge followed by decline.

**Panel 2 — Long-Term Trend** shows the pandemic's fundamental 
momentum after seasonal noise is removed. Two sustained peaks 
are visible — the winter 2020-2021 period and the Delta-Omicron 
period — separated by the spring and summer 2021 trough when 
vaccination was reducing deaths before Delta arrived. 
Importantly the trend never fully returns to zero before the 
analysis period ends in September 2022, consistent with the 
persistent background mortality documented in Hypothesis 2.

**Panel 3 — Seasonal Component** is the most analytically 
important panel for understanding community vulnerability. 
The pattern is clear and consistent across both full years 
visible in the data — peaks in the November through February 
winter window and troughs in the June through August summer 
months. The seasonal component swings approximately 5 deaths 
per 100k above and below the trend — a substantial amplifier 
relative to the overall death rate scale. Winter made COVID 
worse. It did so consistently. And it did so on top of 
whatever structural burden a community was already carrying.

**Panel 4 — Remainder** is largely random noise oscillating 
around zero — which is what a well-specified decomposition 
should produce. The one meaningful exception is the sharp 
positive spike marked with a red dashed line around September 
2021. That is the Delta wave's sudden emergence — more abrupt 
and more severe than the seasonal pattern predicted, producing 
a residual that neither trend nor calendar can explain. Delta 
was a genuinely anomalous event, not simply an amplified 
seasonal peak.

> **The Seasonal Leading Indicator:** The seasonal component 
> begins rising in October — several weeks before the 
> meteorological definition of winter in November. This 
> leading pattern is epidemiologically accurate. People move 
> indoors in early fall as temperatures drop, creating 
> transmission conditions before the calendar catches up. 
> For the communities bearing the highest structural burden 
> this means the dangerous period begins earlier than winter 
> messaging typically acknowledges.


```{r seasonal_by_wave, fig.cap="Monthly death rate by year"}
seasonal_summary <- county_weekly %>%
  mutate(
    month_label = month(week_start, label = TRUE, abbr = TRUE),
    year_fct    = factor(year)
  ) %>%
  group_by(month_label, year_fct) %>%
  summarise(
    death_rate_avg = weighted.mean(death_rate_wk,
                                   total_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(month_label))

max_rate <- max(seasonal_summary$death_rate_avg, na.rm = TRUE)

ggplot(seasonal_summary,
       aes(x     = month_label,
           y     = death_rate_avg,
           color = year_fct,
           group = year_fct)) +

  # Lines and points first
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +

  # Winter shading using geom_tile as background
  # One tile per winter month, full height
  geom_tile(
    data = seasonal_summary %>%
      filter(month_label %in% c("Jan", "Feb", "Nov", "Dec")) %>%
      dplyr::select(month_label) %>%
      distinct() %>%
      mutate(y_mid = max_rate / 2,
             height = max_rate * 1.2),
    aes(x      = month_label,
        y      = y_mid,
        height = height),
    fill        = "#AED6F1",
    alpha       = 0.3,
    inherit.aes = FALSE,
    width       = 1.0
  ) +

  # Redraw lines and points on top of shading
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +

  scale_color_brewer(palette = "Set1", name = "Year") +

  labs(
    title    = "Average Weekly Death Rate by Month and Year",
    subtitle = paste0(
      "Population-weighted national average | ",
      "Blue shading = winter months (Jan, Feb, Nov, Dec) | ",
      "Summer trough near zero in 2021 and 2022"
    ),
    x       = "Month",
    y       = "Avg Weekly Deaths per 100k"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(size = 11)
  )
```

### Seasonal Patterns Across Three Years

The month-by-year chart confirms that the seasonal pattern 
identified in the decomposition is not a mathematical artifact 
— it is visible in the raw monthly averages across all three 
years of the analysis period.

**2021 shows the clearest seasonal shape** — starting high in 
January as the winter 2020-2021 surge peaked, dropping 
steadily through spring and summer to near zero in June and 
July, then rising sharply in September as Delta emerged 
simultaneously with autumn. The summer floor and autumn rise 
are exactly what the seasonal component predicted.

**2022 shows the post-Omicron floor** — the lowest death rates 
of any year across almost every month, consistent with 
Hypothesis 2's finding of structural decline in case lethality. 
But the seasonal shape persists — near zero in summer, rising 
in fall — confirming that seasonality continued even as the 
overall mortality level declined dramatically.

**2020 is the anomalous year** — the pandemic arrived in the 
wrong season. The April peak reflects the initial uncontrolled 
outbreak hitting a population with zero immunity and zero 
treatment protocols before lockdowns suppressed transmission 
through late spring. The 2020 line maps almost precisely onto 
the policy decisions of that year — lockdowns, reopenings, 
and the arrival of winter without adequate preparation — 
rather than the biological seasonal pattern that 2021 and 
2022 would establish.

> **Seasonal Equity Note:** The consistent summer trough — 
> near zero in both 2021 and 2022 regardless of variant or 
> vaccination status — suggests that summer provided genuine 
> biological relief to all communities. The communities 
> documented in this analysis as persistently high burden 
> did not escape that summer relief. But they entered each 
> autumn with higher cumulative burden, less healthcare 
> infrastructure, and fewer resources to weather the winter 
> amplification that consistently followed. Seasonality was 
> equal. Its consequences were not.

# 9. Outlier County Detection

```{r funnel_plot, fig.cap="Funnel plot: county death rate vs population size"}
library(plotly)

# ============================================================
# Recalculate funnel limits and county flags
# ============================================================
overall_rate <- sum(county_totals$total_deceased, na.rm = TRUE) /
                sum(county_totals$total_pop,      na.rm = TRUE)

pop_range <- seq(
  min(county_totals$total_pop, na.rm = TRUE),
  max(county_totals$total_pop, na.rm = TRUE),
  length.out = 500
)

funnel_limits <- tibble(
  total_pop = pop_range,
  se        = sqrt(overall_rate * (1 - overall_rate) / pop_range),
  ucl_95    = (overall_rate + 1.96 * se) * 100000,
  lcl_95    = (overall_rate - 1.96 * se) * 100000,
  ucl_998   = (overall_rate + 3.09 * se) * 100000,
  lcl_998   = (overall_rate - 3.09 * se) * 100000
)

county_totals <- county_totals %>%
  mutate(
    se_county    = sqrt(overall_rate * (1 - overall_rate) / total_pop),
    ucl_998      = (overall_rate + 3.09 * se_county) * 100000,
    lcl_998      = (overall_rate - 3.09 * se_county) * 100000,
    funnel_flag  = case_when(
      cum_death_rate > ucl_998 ~ "Above 99.8% limit",
      cum_death_rate < lcl_998 ~ "Below 99.8% limit",
      TRUE                     ~ "Within limits"
    ),
    expected_rate = overall_rate * 100000,
    excess_deaths = round(
      (cum_death_rate - overall_rate * 100000) / 100000 * total_pop, 0
    ),

    # Full tooltip for every county
    tooltip_text = paste0(
      "<b>", county_name, ", ", state_name, "</b><br>",
      "────────────────────────<br>",
      "<b>Funnel Status: ", funnel_flag, "</b><br>",
      "────────────────────────<br>",
      "Cumulative Death Rate: ",
      round(cum_death_rate, 1), " per 100k<br>",
      "National Expected Rate: ",
      round(overall_rate * 100000, 1), " per 100k<br>",
      "Difference from Expected: ",
      round(cum_death_rate - overall_rate * 100000, 1),
      " per 100k<br>",
      "Estimated Excess Deaths: ",
      scales::comma(abs(excess_deaths)),
      if_else(excess_deaths >= 0,
              " above expected<br>",
              " below expected<br>"),
      "────────────────────────<br>",
      "Population: ", scales::comma(total_pop), "<br>",
      "% Age 65+: ",
      scales::percent(pct_65_over,  accuracy = 0.1), "<br>",
      "% Black: ",
      scales::percent(pct_black,    accuracy = 0.1), "<br>",
      "Poverty Rate: ",
      scales::percent(poverty_rate, accuracy = 0.1), "<br>",
      "% Without HS Diploma: ",
      scales::percent(pct_less_than_hs, accuracy = 0.1), "<br>",
      "────────────────────────<br>",
      "<i>Upper 99.8% limit at this population: ",
      round(ucl_998, 1), " per 100k</i>"
    )
  )

# ============================================================
# Build base ggplot
# ============================================================
p_funnel <- ggplot() +

  # Control limit ribbons
  geom_ribbon(data = funnel_limits,
              aes(x = total_pop,
                  ymin = lcl_998, ymax = ucl_998),
              fill = "#457B9D", alpha = 0.12,
              inherit.aes = FALSE) +
  geom_ribbon(data = funnel_limits,
              aes(x = total_pop,
                  ymin = lcl_95, ymax = ucl_95),
              fill = "#457B9D", alpha = 0.20,
              inherit.aes = FALSE) +

  # National mean reference line
  geom_hline(
    yintercept = overall_rate * 100000,
    linetype   = "dashed",
    color      = "grey40",
    linewidth  = 0.7
  ) +

  # County points with tooltip
  geom_point(
    data = county_totals,
    aes(x    = total_pop,
        y    = cum_death_rate,
        color = funnel_flag,
        text  = tooltip_text),
    alpha = 0.6,
    size  = 1.8
  ) +

  scale_color_manual(
    values = c(
      "Above 99.8% limit" = "#E63946",
      "Below 99.8% limit" = "#2A9D8F",
      "Within limits"     = "grey60"
    ),
    name = NULL
  ) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +

  labs(
    title    = "Funnel Plot: County COVID-19 Death Rate vs. Population",
    subtitle = paste0(
      "Hover any county for demographic profile | ",
      "Red = statistically anomalous high burden | ",
      "Teal = anomalous low burden"
    ),
    x       = "County Population (log scale)",
    y       = "Cumulative Deaths per 100k",
    caption = paste0(
      "Shaded bands = 95% and 99.8% control limits | ",
      "Dashed line = national mean (",
      round(overall_rate * 100000, 1), " per 100k) | ",
      "Counties outside 99.8% band are statistically ",
      "anomalous for their population size"
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

# ============================================================
# Convert to interactive plotly
# ============================================================
ggplotly(p_funnel, tooltip = "text") %>%
  layout(
    hoverlabel = list(
      bgcolor     = "white",
      font        = list(size = 11),
      bordercolor = "grey40",
      align       = "left"
    ),
    legend = list(orientation = "h", y = -0.15)
  )
```



# 10. Wave-Stratified Analysis

```{r wave_death_rates, fig.cap="County death rate distribution by pandemic wave"}
# Clean wave labels for non-technical audience
wave_labels <- c(
  "1_initial"    = "Wave 1: Initial Outbreak\n(Mar-Jun 2020)",
  "2_summer2020" = "Wave 2: Summer 2020\n(Jul-Sep 2020)",
  "3_winter2021" = "Wave 3: Winter Surge\n(Oct 2020-Feb 2021)",
  "4_alpha"      = "Wave 4: Alpha Variant\n(Mar-Jun 2021)",
  "5_delta"      = "Wave 5: Delta Variant\n(Jul-Nov 2021)",
  "6_omicron"    = "Wave 6: Omicron\n(Dec 2021-Mar 2022)",
  "7_endemic"    = "Wave 7: Endemic Phase\n(Apr-Sep 2022)"
)

# Calculate wave medians for annotation
wave_medians <- county_weekly %>%
  filter(pandemic_wave != "pre_pandemic") %>%
  group_by(fips, pandemic_wave, total_pop) %>%
  summarise(
    wave_death_rate = (sum(new_deceased_wk, na.rm = TRUE) /
                       first(total_pop)) * 100000,
    .groups = "drop"
  ) %>%
  filter(wave_death_rate >= 0) %>%
  group_by(pandemic_wave) %>%
  summarise(
    median_dr = median(wave_death_rate, na.rm = TRUE),
    p75_dr    = quantile(wave_death_rate, 0.75, na.rm = TRUE),
    .groups   = "drop"
  )

county_weekly %>%
  filter(pandemic_wave != "pre_pandemic") %>%
  group_by(fips, county_name, state_name,
           pandemic_wave, total_pop) %>%
  summarise(
    wave_death_rate = (sum(new_deceased_wk, na.rm = TRUE) /
                       first(total_pop)) * 100000,
    .groups = "drop"
  ) %>%
  filter(wave_death_rate >= 0) %>%
  # Trim to 95th percentile to remove extreme outliers
  # that compress the meaningful distribution
  filter(wave_death_rate <=
           quantile(wave_death_rate, 0.95, na.rm = TRUE)) %>%
  mutate(
    pandemic_wave = factor(
      pandemic_wave,
      levels  = names(wave_labels),
      labels  = wave_labels
    )
  ) %>%
  ggplot(aes(x    = wave_death_rate,
             y    = pandemic_wave,
             fill = pandemic_wave)) +
  ggridges::geom_density_ridges(
    alpha          = 0.8,
    scale          = 1.4,
    quantile_lines = TRUE,
    quantiles      = 2,          # median line
    linewidth      = 0.6
  ) +
  scale_fill_manual(
  values = c(
    "Wave 1: Initial Outbreak\n(Mar-Jun 2020)"       = "#E63946",
    "Wave 2: Summer 2020\n(Jul-Sep 2020)"            = "#F4A261",
    "Wave 3: Winter Surge\n(Oct 2020-Feb 2021)"      = "#457B9D",
    "Wave 4: Alpha Variant\n(Mar-Jun 2021)"          = "#2A9D8F",
    "Wave 5: Delta Variant\n(Jul-Nov 2021)"          = "#E76F51",
    "Wave 6: Omicron\n(Dec 2021-Mar 2022)"           = "#6A0572",
    "Wave 7: Endemic Phase\n(Apr-Sep 2022)"          = "#A8DADC"
  )
) + 
  scale_x_continuous(
    labels = comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.02))
  ) +
  # Median value labels on each ridge
  geom_text(
    data = wave_medians %>%
      mutate(
        pandemic_wave = factor(
          pandemic_wave,
          levels = names(wave_labels),
          labels = wave_labels
        )
      ),
    aes(x     = median_dr + 4,
        y     = pandemic_wave,
        label = paste0("Median: ",
                       round(median_dr, 1))),
    inherit.aes = FALSE,
    hjust       = 0,
    vjust       = -0.8,
    size        = 3.0,
    color       = "grey30",
    fontface    = "bold"
  ) +
  labs(
    title    = "County Death Rate Distribution by Pandemic Wave",
    subtitle = paste0(
      "Each ridge = distribution of county death rates during that wave | ",
      "Vertical line = median | ",
      "Trimmed at 95th percentile for readability"
    ),
    x       = "Deaths per 100k during wave period",
    y       = NULL,
    caption = paste0(
      "Wave periods defined by dominant variant circulation | ",
      "Extreme outlier counties (top 5%) excluded from display | ",
      "All counties included in statistical analysis"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "none",
    axis.text.y      = element_text(size = 9, hjust = 1),
    plot.caption     = element_text(size = 8, color = "grey50"),
    panel.grid.minor = element_blank()
  )
```


# 11. Next Steps & Data Gaps

## Data Limitations & Coverage Assessment

**Coverage Gap: covid19_open_data Pipeline**

Six states have zero county-level coverage in the Google 
covid19_open_data pipeline: Arkansas, Alabama, Colorado, 
California, Arizona, and Connecticut (287 counties total). 
An additional 28 states have partial coverage with Texas 
missing 50 of 254 counties and several Great Plains states 
missing 30-40% of counties.

Demographic comparison of included vs missing counties reveals 
the coverage gap is less directionally biased than initially 
characterized. Missing counties (n=285) have substantially 
smaller populations (median 2,860 vs 30,463), lower Black 
population shares (2.5% vs 9.1%), lower poverty rates 
(13.1% vs 15.5%), and higher elderly population shares 
(23.0% vs 18.4%) than included counties. This profile is 
consistent with small rural counties in the Great Plains and 
Mountain West rather than the high-poverty majority minority 
communities driving the core structural inequity findings.

The Deep South concentration finding is therefore unlikely 
to be materially biased by the coverage gap. The primary 
residual concern is Alabama's complete absence -- a state 
with significant Black Belt geography that almost certainly 
contains counties meeting the persistently high burden 
criteria. Core findings are assessed as robust to the current 
data gap but complete national coverage remains a Phase 2 
priority.

**Time Series Gap: September 2022 to May 2023**

The covid19_open_data pipeline stopped updating in September 
2022. The US Public Health Emergency did not officially end 
until May 11, 2023. The full Omicron tail and the transition 
to endemic surveillance are not captured in the current 
analysis period.



## Phase 2 Data Integration Plan

**Priority 1 — Complete National Coverage**
USAFacts confirmed cases and deaths data covers all 3,143 US 
counties and will fill the state-level pipeline gaps after 
resolving the wide-format date unpivoting challenge documented 
in the original ETL. CDC Wonder mortality data provides an 
alternative complete-coverage source joinable on FIPS. Either 
source will extend the analysis to full national coverage and 
remove the systematic geographic bias documented above.

**Priority 2 — CDC PLACES Comorbidity Data**
County-level crude prevalence for diabetes, obesity, COPD, 
coronary heart disease, hypertension, and smoking. Joins on 
FIPS. Will replace the ACS disability proxy removed from the 
ETL and give the regression model proper clinical comorbidity 
covariates — the single most important analytical gap in the 
current model specification.

**Priority 3 — CDC Vaccination Rate by County**
County-level vaccination coverage data joins on FIPS and will 
allow the regression model to formally partition variant 
severity from vaccination protection effects during the Delta 
wave. This directly addresses the H1 hypothesis limitation 
noted in Section 3b — the vaccination covariate needed to 
formally test whether uptake differences explain the Wave 1 
to Delta burden reorganization.

**Priority 4 — CMS Medicaid and Medicare Integration**
The persistently high burden counties identified in H3 map 
closely onto regions of historically elevated Medicaid 
enrollment and constrained healthcare infrastructure. CMS 
publishes county-level Medicaid enrollment and per-capita 
spending data joinable on FIPS. Formal testing of whether 
healthcare access predicted persistent burden independent of 
demographic covariates is the most direct policy-actionable 
extension of the current findings.

**Priority 5 — HRSA Health Professional Shortage Areas**
County-level federal designations of medically underserved 
areas join on FIPS and will test whether the persistently 
high burden counties were already identified as healthcare 
access deserts before the pandemic arrived. The expected 
overlap between HRSA shortage designations and the 157 
persistently high burden counties is a direct measure of 
the degree to which COVID mortality was predictable from 
pre-existing structural conditions.

**Priority 6 — Native American Community Analysis**
McKinley County New Mexico (majority Navajo Nation) appears 
in the top 30 extreme mortality counties with a death rate 
of 813.1 per 100k despite an elderly population of just 
11.9%. Indian Health Service coverage data and tribal census 
demographics are available at the county and tribal level 
and would allow formal comparison of Native American 
community outcomes against the Deep South Black Belt and 
Appalachian poverty patterns identified in this analysis. 
All three represent distinct histories of structural federal 
neglect with convergent pandemic outcomes.



## Phase 2 Modeling Extensions

**Mixed-Effects Model**
The current negative binomial model treats counties as 
independent observations. A mixed-effects extension using 
lme4::glmer.nb with state-level random effects would account 
for the fact that counties within the same state share policy 
environments, reporting standards, and healthcare 
infrastructure. This is particularly important given the 
strong state-level clustering visible in the persistently 
high burden county geographic distribution.

**Time-Varying Covariate Model**
The current model uses static ACS demographic covariates. 
A panel model incorporating time-varying vaccination rates, 
mobility indices, and variant-specific wave indicators would 
allow formal testing of whether structural disadvantage 
interacted with pandemic phase to produce the wave 
persistence pattern documented in H3.

**Facility-Level Outlier Analysis**
aggregation_level = 3 in covid19_open_data contains 
sub-county localities including nursing homes and care 
facilities in some states. Facility-level analysis would 
identify whether specific institutions drove county-level 
outlier status — particularly relevant for the extreme 
mortality counties where a single nursing home outbreak 
in a small population could produce catastrophic rates.



## Phase 2 Visualization Plan

**Interactive National Choropleth (Leaflet)**
The current static choropleth will be rebuilt in Phase 2 
using leaflet with complete national coverage. The 
interactive version will include state boundary overlays, 
county-level demographic tooltips, a state filter dropdown 
for targeted analysis, and wave-specific views allowing 
the geographic reorganization documented in H3 to be 
explored visually at the county level.

**Persistently High Burden County Profile Cards**
The 157 persistently high burden counties currently 
represented in the interactive leaflet map will be 
supplemented with Phase 2 data to include Medicaid 
enrollment rates, HRSA shortage designations, CDC PLACES 
chronic disease prevalence, and vaccination coverage. 
Each county's tooltip will become a complete structural 
vulnerability profile.



*Document generated: `r Sys.Date()`*
*Data: Google COVID-19 Open Data + US Census ACS 2019 5-Year*
*Analysis period: March 2020 -- September 2022*
*Counties analyzed: 2,904 of approximately 3,143 US counties (92%)*
*Phase 2 data integration in progress*


*Document generated: `r Sys.Date()`*  
*Data: Google COVID-19 Open Data + US Census ACS 2019 5-Year*  
*Analysis period: March 2020 - September 2022*
